# Local Tool Installation for All Practicals

This R Markdown guide installs a shared local tool stack for the whole workspace.

The installer writes everything under:

- `tools/`

This shared folder can then be used by `practical2`, `practical3`, and `practical4`.

The installer downloads or prepares these tools:

- `popoolation2`
- `npstat`
- `structure`
- `CLUMPP`
- `structureHarvester`

It also creates a local runtime environment under `tools/env` and wrapper commands under `tools/bin`.

`structure` is installed from the Bioconda package `bioconda::structure` into `tools/env`.

## Setup

Run this first chunk to point R to the workspace root.

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
repo_root <- normalizePath(getwd(), winslash = "/", mustWork = TRUE)
installer <- file.path(repo_root, "install_local_tools.sh")
tools_dir <- file.path(repo_root, "tools")
bin_dir <- file.path(tools_dir, "bin")
env_lib_dir <- file.path(tools_dir, "env", "lib")
perl5_dir <- file.path(tools_dir, "perl5", "lib", "perl5")
```

If you opened this file from a subfolder project instead of the workspace root, reset `repo_root` manually before running the installer.

```{r, eval=FALSE}
repo_root <- "/home/rstudio/work/Evolutionary-Genomics-UNIFR-2026-SBL.20034"
installer <- file.path(repo_root, "install_local_tools.sh")
tools_dir <- file.path(repo_root, "tools")
bin_dir <- file.path(tools_dir, "bin")
env_lib_dir <- file.path(tools_dir, "env", "lib")
perl5_dir <- file.path(tools_dir, "perl5", "lib", "perl5")
```

## Run the Installer

This calls the shared installer at the workspace root.

```{r install-tools, eval=FALSE}
stopifnot(file.exists(installer))
system2("bash", installer)
```

## Expose the Tools in RStudio

After installation, set the runtime environment for the current R session.

```{r set-env, eval=FALSE}
Sys.setenv(
  PATH = paste(bin_dir, Sys.getenv("PATH"), sep = ":"),
  LD_LIBRARY_PATH = paste(env_lib_dir, Sys.getenv("LD_LIBRARY_PATH"), sep = ":"),
  PERL5LIB = paste(perl5_dir, Sys.getenv("PERL5LIB"), sep = ":")
)
```

## Quick Checks

Use these commands to verify that the wrappers are visible from the R session.

```{r verify-tools, eval=FALSE}
system("which npstat")
system("which structure")
system("which CLUMPP")
system("which structureHarvester")
system("which fst-sliding.pl")
```

You can also inspect the installed tool tree.

```{r inspect-files, eval=FALSE}
list.files(tools_dir)
list.files(bin_dir)
```

## Important Note About STRUCTURE

`structure` is installed from Bioconda into the shared local environment.

The wrapper in `tools/bin/structure` runs the executable from `tools/env/bin/structure`.

So:

- `npstat`, `popoolation2`, `CLUMPP`, `structureHarvester`, and `structure` are all managed from the shared local setup

If `structure` fails, check whether the local micromamba environment finished installing successfully.

## Example From a Practical Rmd

If you want to call the shared installer from another practical notebook, the pattern is:

```{r example-from-practical, eval=FALSE}
system2("bash", "../install_local_tools.sh")
Sys.setenv(
  PATH = paste("../tools/bin", Sys.getenv("PATH"), sep = ":")
)
```