---
title: "Practical 3 Part 1: STRUCTURE from ddRAD SNPs"
output: html_document
---

## Hide setup details

This first chunk only sets global notebook options. Students do not need to edit it.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

Load the R packages used for plotting, reshaping the ancestry table, and drawing the population map.

```{r libraries}
library(cowplot)
library(dplyr)
library(ggplot2)
library(reshape2)
library(rworldmap)
```

## Set the practical root directory

Move to the scripts directory before reading inputs or writing outputs.

```{r working-directory}
setwd("/home/rstudio/work/Evolutionary-Genomics-UNIFR-2026-SBL.20034/practical3/scripts")
getwd()
```

## Set output and input paths

Define where results will be written and which precomputed files will be used in this part of the practical.

```{r options}
output_dir <- "/home/rstudio/work/outputs/01-structure-ddrad"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

evanno_file <- file.path(output_dir, "structureHarvester_analysis", "evanno.txt")
sample_metadata_file <- "/home/rstudio/work/dataset-tutorial3/01-structure-ddrad/04-sample-metadata/samples_xy.txt"
```

## Read the model-choice summary

This chunk reads the Evanno summary table across values of `K`. These are the numbers you will use to decide which `K` is most strongly supported.

```{r read-evanno-summary}
evanno <- read.table(evanno_file, comment.char = "#", header = FALSE, stringsAsFactors = FALSE)
colnames(evanno) <- c("k", "replicates", "mean_ln_prob", "sd_ln_prob", "ln_prime", "ln_double_prime", "delta_k")

evanno$k <- as.numeric(evanno$k)
evanno$mean_ln_prob <- as.numeric(evanno$mean_ln_prob)
evanno$sd_ln_prob <- as.numeric(evanno$sd_ln_prob)
evanno$delta_k <- as.numeric(evanno$delta_k)

write.table(
        evanno,
        file.path(output_dir, "structure_evanno_summary.tsv"),
        sep = "\t",
        quote = FALSE,
        row.names = FALSE
)

evanno
```

## Plot model choice

Look at these two panels before doing anything else. The first shows how the mean log probability changes with `K`, and the second shows the Delta K criterion used to identify the most likely number of clusters.

```{r plot-model-choice, fig.width=8, fig.height=7}
ln_prob_plot <- ggplot(evanno, aes(x = k, y = mean_ln_prob)) +
        geom_line(linewidth = 0.6, colour = "#264653") +
        geom_point(size = 2.3, colour = "#264653") +
        geom_errorbar(
                aes(ymin = mean_ln_prob - sd_ln_prob, ymax = mean_ln_prob + sd_ln_prob),
                width = 0.15,
                linewidth = 0.45,
                colour = "#264653"
        ) +
        scale_x_continuous(breaks = evanno$k) +
        labs(title = "How support changes across K", x = "K", y = "Mean ln P(D)") +
        theme_minimal(base_size = 11)

delta_plot <- ggplot(evanno, aes(x = k, y = delta_k)) +
        geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey70") +
        geom_col(width = 0.75, fill = "#E76F51", na.rm = TRUE) +
        scale_x_continuous(breaks = evanno$k) +
        labs(title = "Use this panel to choose the best K", x = "K", y = "Delta K") +
        theme_minimal(base_size = 11)

model_choice_plot <- cowplot::plot_grid(
        ln_prob_plot,
        delta_plot,
        ncol = 1,
        align = "v",
        rel_heights = c(1.1, 1)
)

ggsave(
        file.path(output_dir, "structure_model_choice.png"),
        model_choice_plot,
        width = 8,
        height = 7,
        dpi = 320
)

model_choice_plot
```

## Choose the best K

After inspecting the model-choice figure, replace the value below with the `K` you think best summarizes the data. Keep this as a deliberate student decision.

```{r choose-best-k}
# After looking at the model-choice figure above, replace 4 with the K you think is best.
best_k <- 8 # <- REPLACE THIS NUMBER WITH YOUR CHOICE

best_k
```

## Read sample coordinates

Load the table that contains the individual names, population labels, and map coordinates used for the final plots.

```{r read-sample-metadata}
samples_xy <- read.table(sample_metadata_file, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
head(samples_xy)
```

## Read ancestry coefficients for the chosen K

Now read the aligned CLUMPP membership file that matches your chosen `K`. This creates one ancestry coefficient column per inferred cluster.

```{r read-membership-table}
membership_file <- file.path(output_dir, "clumpp", paste0("clumpp_output_", best_k, ".txt"))

membership_raw <- read.table(membership_file, header = FALSE, stringsAsFactors = FALSE)
cluster_columns <- paste0("cluster_", seq_len(best_k))
cluster_colours <- hcl.colors(best_k, palette = "Set 2")

membership_values <- membership_raw[, (ncol(membership_raw) - best_k + 1):ncol(membership_raw)]
colnames(membership_values) <- cluster_columns

membership <- cbind(
        ind = samples_xy$Ind,
        X = samples_xy$X,
        Y = samples_xy$Y,
        pop = samples_xy$pop,
        membership_values
)

write.table(
        membership,
        file.path(output_dir, paste0("structure_k", best_k, "_membership.tsv")),
        sep = "\t",
        quote = FALSE,
        row.names = FALSE
)

head(membership)
```

## Reshape the ancestry table

Convert the wide membership table into long format so it can be drawn easily as a stacked ancestry barplot.

```{r reshape-membership-for-plot}
membership_long <- reshape2::melt(membership[, c("ind", cluster_columns)], id.vars = "ind")
colnames(membership_long) <- c("ind", "cluster", "value")
membership_long$ind <- factor(membership_long$ind, levels = membership$ind)

head(membership_long)
```

## Plot individual ancestry

Each bar is one individual. The colored fractions show the estimated ancestry membership in each cluster for your chosen `K`.

```{r ancestry-barplot, fig.width=12, fig.height=5}
ancestry_barplot <- ggplot(membership_long, aes(x = ind, y = value, fill = cluster)) +
        geom_bar(position = "fill", stat = "identity", width = 0.8) +
        theme_minimal(base_size = 11) +
        theme(axis.text.x = element_text(size = 8, angle = 90, hjust = 0.8, vjust = 0.5)) +
        scale_fill_manual(values = cluster_colours) +
        labs(
                title = paste("Individual ancestry proportions for K =", best_k),
                x = NULL,
                y = "Membership coefficient",
                fill = "Cluster"
        )

ggsave(
        file.path(output_dir, paste0("structure_k", best_k, "_barplot.png")),
        ancestry_barplot,
        width = 12,
        height = 5,
        dpi = 320
)

ancestry_barplot
```

## Calculate population means

Average the individual ancestry coefficients within each population. These mean values are used to build the population map.

```{r population-means}
population_means <- membership[, c("X", "Y", "pop", cluster_columns)] |>
        dplyr::group_by(pop) |>
        dplyr::summarise(dplyr::across(c("X", "Y", dplyr::all_of(cluster_columns)), mean), .groups = "drop")

population_means
```

## Draw the population map

This chunk places each population on the Switzerland map and adds a pie chart showing the mean ancestry proportions at that location.

```{r population-map}
map_x_limits <- c(min(population_means$X) - 2, max(population_means$X) + 2)
map_y_limits <- c(min(population_means$Y) - 2, max(population_means$Y) + 2)

png(file.path(output_dir, "population_sampling_map.png"), width = 8, height = 5, units = "in", res = 320)
plot(
        rworldmap::getMap(resolution = "low"),
        xlim = map_x_limits,
        ylim = map_y_limits,
        asp = 1,
        col = "white",
        border = "grey65",
        bg = "aliceblue"
)
rworldmap::mapPies(
        population_means,
        nameX = "X",
        nameY = "Y",
        symbolSize = 1,
        nameZs = cluster_columns,
        zColours = cluster_colours,
        xlim = map_x_limits,
        ylim = map_y_limits,
        oceanCol = "aliceblue",
        landCol = "white",
        addCatLegend = FALSE
)
text(population_means$X, population_means$Y + 0.35, labels = population_means$pop, cex = 0.85, font = 2)
title(main = paste("Population coordinates and mean admixture proportions for K =", best_k))
dev.off()
```
