---
title: "Practical 3 Part 2: ddRAD and WGS comparison"
output: html_document
---

## Hide setup details


This first chunk only sets global notebook options. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
knitr::opts_knit$set(root.dir = "/home/rstudio/work")
```

## Load packages

Load the R packages needed for PCA, diversity summaries, Mantel tests, and plotting.

```{r libraries}
library(adegenet)
library(cowplot)
library(dplyr)
library(ggplot2)
library(hierfstat)
library(tidyr)
library(vcfR)
library(vegan)
```

## Set plot theme and paths

Define the shared plotting style, the output folder, and the population order used throughout the analysis.

```{r options}
theme_set(
        theme_minimal(base_size = 11) +
                theme(
                        plot.title.position = "plot",
                        legend.position = "bottom",
                        panel.grid.minor = element_blank()
                )
)

output_dir <- "/home/rstudio/work/outputs/02-ddrad-wgs-population-comparison"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

population_order <- c("A2G", "A2S", "V2B", "B2MV", "P2", "K2S2")
population_colours <- c(
        A2G = "#1B9E77",
        A2S = "#D95F02",
        V2B = "#7570B3",
        B2MV = "#E7298A",
        P2 = "#66A61E",
        K2S2 = "#E6AB02"
)
```

## Read the input data

Load the elevation table, the geographic distance matrix, and the ddRAD and WGS VCF files.

```{r read-inputs}
elevation_table <- read.table(
        "/home/rstudio/work/dataset-tutorial3/02-ddrad-wgs-population-comparison/elevation.txt",
        header = FALSE,
        stringsAsFactors = FALSE
)
names(elevation_table) <- c("pop", "elevation_m")

geo_table <- read.table(
        "/home/rstudio/work/dataset-tutorial3/02-ddrad-wgs-population-comparison/geo_dist.txt",
        header = TRUE,
        check.names = FALSE
)
geo_matrix <- as.matrix(geo_table)

ddrad_vcf <- read.vcfR(
        "/home/rstudio/work/dataset-tutorial3/02-ddrad-wgs-population-comparison/ddRAD_filtered_pruned.vcf.gz",
        verbose = FALSE
)

wgs_vcf <- read.vcfR(
        "/home/rstudio/work/dataset-tutorial3/02-ddrad-wgs-population-comparison/diploids_bisnps_filt1rem_minDP4nc_maxDP1009rm_filtMAC4MAF01_MD02_pruned.vcf.gz",
        verbose = FALSE
)
```

## Prepare ddRAD metadata

Extract sample names and population labels from the ddRAD VCF, then attach elevation information for plotting.

```{r ddrad-metadata}
ddrad_genlight <- vcfR2genlight(ddrad_vcf)
ddrad_sample_ids <- indNames(ddrad_genlight)
ddrad_populations <- sub("-.*", "", ddrad_sample_ids)
ddrad_populations[ddrad_populations == "s2"] <- "K2S2"

ddrad_metadata <- data.frame(
        sample_id = ddrad_sample_ids,
        pop = ddrad_populations,
        stringsAsFactors = FALSE
)
ddrad_metadata$elevation_m <- elevation_table$elevation_m[match(ddrad_metadata$pop, elevation_table$pop)]
ddrad_metadata$pop <- factor(ddrad_metadata$pop, levels = population_order)
ddrad_metadata$elevation_band <- ifelse(ddrad_metadata$elevation_m >= 1000, "High", "Low")

locNames(ddrad_genlight) <- paste(ddrad_vcf@fix[, 1], ddrad_vcf@fix[, 2], sep = "_")
pop(ddrad_genlight) <- ddrad_metadata$pop
```

## Run ddRAD PCA

Calculate the ddRAD PCA, save the explained variance table, and compute the coordinates needed for later PCA plots.

```{r ddrad-pca}
ddrad_pca <- glPca(ddrad_genlight, nf = 4)

ddrad_explained_variance <- data.frame(
        principal_component = seq_along(ddrad_pca$eig),
        explained_variance_percent = round(100 * ddrad_pca$eig / sum(ddrad_pca$eig), 2)
)
ddrad_explained_variance$principal_component_label <- paste0("PC", ddrad_explained_variance$principal_component)

ddrad_scores <- as.data.frame(ddrad_pca$scores[, 1:4, drop = FALSE])
ddrad_scores$sample_id <- ddrad_metadata$sample_id
ddrad_scores$pop <- ddrad_metadata$pop
ddrad_scores$elevation_band <- ddrad_metadata$elevation_band

ddrad_ellipse_scores <- ddrad_scores |>
        dplyr::filter(is.finite(PC1), is.finite(PC2), !is.na(pop)) |>
        dplyr::group_by(pop) |>
        dplyr::filter(dplyr::n() >= 3) |>
        dplyr::ungroup()

write.table(
        ddrad_explained_variance,
        file = file.path(output_dir, "ddrad_pca_explained_variance.tsv"),
        sep = "\t",
        quote = FALSE,
        row.names = FALSE
)

ddrad_explained_variance
```

## Plot ddRAD eigenvalues

This scree-style plot shows how much variation is explained by each principal component in the ddRAD dataset.

Question: Do the first one or two ddRAD axes capture most of the variation, or is the signal spread across several components?

```{r ddrad-eigenvalues, fig.width=6.5, fig.height=4.5}
ddrad_eigenvalue_data <- as.data.frame(ddrad_explained_variance)
ddrad_eigenvalue_data$principal_component_num <- suppressWarnings(as.integer(sub("^PC", "", ddrad_eigenvalue_data$principal_component)))
if (any(is.na(ddrad_eigenvalue_data$principal_component_num))) {
        ddrad_eigenvalue_data$principal_component_num <- seq_len(nrow(ddrad_eigenvalue_data))
}
ddrad_eigenvalue_data$principal_component_label <- paste0("PC", ddrad_eigenvalue_data$principal_component_num)

ddrad_eigenvalue_plot <- ggplot(ddrad_eigenvalue_data, aes(x = principal_component_num, y = explained_variance_percent)) +
        geom_col(fill = "#1B9E77", width = 0.75) +
        geom_text(aes(label = sprintf("%.1f", explained_variance_percent)), vjust = -0.3, size = 3) +
        scale_x_continuous(
                breaks = ddrad_eigenvalue_data$principal_component_num[seq(1, nrow(ddrad_eigenvalue_data), by = 5)],
                labels = ddrad_eigenvalue_data$principal_component_label[seq(1, nrow(ddrad_eigenvalue_data), by = 5)]
        ) +
        labs(title = "ddRAD PCA eigenvalues", x = "Principal component", y = "Explained variance (%)") +
        theme_minimal(base_size = 11)

ggsave(file.path(output_dir, "ddrad_pca_eigenvalues.png"), ddrad_eigenvalue_plot, width = 6.5, height = 4.5, dpi = 320)
ddrad_eigenvalue_plot
```

## Plot ddRAD loadings

This plot shows the full distribution of ddRAD marker loadings across the first four PCA axes. Dense horizontal bands near zero mean that most markers contribute weakly, whereas isolated vertical spikes indicate a smaller number of especially influential markers. The exported top-loading table lists the 20 markers with the strongest absolute contributions on the first two axes.

Question: Do the first ddRAD axes look driven by a few concentrated loading peaks, or by many markers with modest effects?

```{r ddrad-loadings, fig.width=9, fig.height=6}
ddrad_loading_matrix <- as.data.frame(ddrad_pca$loadings[, 1:4, drop = FALSE])
colnames(ddrad_loading_matrix) <- paste0("PC", seq_len(ncol(ddrad_loading_matrix)))
ddrad_loading_matrix$marker <- rownames(ddrad_loading_matrix)
ddrad_loading_matrix$marker_index <- seq_len(nrow(ddrad_loading_matrix))

ddrad_loading_table <- tidyr::pivot_longer(
        ddrad_loading_matrix[, c("marker", "PC1", "PC2")],
        cols = c("PC1", "PC2"),
        names_to = "axis",
        values_to = "loading"
)
ddrad_loading_table$abs_loading <- abs(ddrad_loading_table$loading)
ddrad_loading_table <- ddrad_loading_table |>
        dplyr::group_by(axis) |>
        dplyr::slice_max(order_by = abs_loading, n = 20, with_ties = FALSE) |>
        dplyr::ungroup() |>
        dplyr::arrange(axis, dplyr::desc(abs_loading))

ddrad_loading_plot_data <- tidyr::pivot_longer(
        ddrad_loading_matrix[, c("marker_index", "PC1", "PC2", "PC3", "PC4")],
        cols = c("PC1", "PC2", "PC3", "PC4"),
        names_to = "axis",
        values_to = "loading"
)

write.table(
        ddrad_loading_table,
        file = file.path(output_dir, "ddrad_pca_top_loadings.tsv"),
        sep = "\t",
        quote = FALSE,
        row.names = FALSE
)

ddrad_loading_plot <- ggplot(ddrad_loading_plot_data, aes(x = marker_index, y = loading)) +
        geom_bin2d(bins = 70) +
        facet_wrap(~axis, scales = "free_y") +
        scale_fill_viridis_c(option = "C", name = "Count") +
        labs(title = "ddRAD PCA loading distributions", x = "Marker index", y = "Loading") +
        theme_minimal(base_size = 11)

ggsave(file.path(output_dir, "ddrad_pca_top_loadings.png"), ddrad_loading_plot, width = 9, height = 6, dpi = 320)
ddrad_loading_plot
```

## Plot ddRAD PCA

These four panels show how individuals cluster in ddRAD space across several pairs of principal components. Looking beyond PC1 and PC2 helps distinguish whether separation is concentrated on the first axis pair or also appears on later axes.

```{r ddrad-pca-plot, fig.width=10, fig.height=8}
ddrad_pca_plot_pc12 <- ggplot(ddrad_scores, aes(x = PC1, y = PC2, colour = pop, shape = elevation_band)) +
        stat_ellipse(data = ddrad_ellipse_scores, aes(x = PC1, y = PC2, group = pop, colour = pop), type = "norm", linewidth = 0.4, alpha = 0.15, show.legend = FALSE) +
        geom_point(size = 2.1, alpha = 0.9) +
        scale_colour_manual(values = population_colours, drop = FALSE) +
        labs(
                title = "PC1 vs PC2",
                x = paste0("PC1 (", ddrad_explained_variance$explained_variance_percent[1], "%)"),
                y = paste0("PC2 (", ddrad_explained_variance$explained_variance_percent[2], "%)"),
                colour = "Population",
                shape = "Elevation"
        )

ddrad_pca_plot_pc13 <- ggplot(ddrad_scores, aes(x = PC1, y = PC3, colour = pop, shape = elevation_band)) +
        stat_ellipse(data = ddrad_ellipse_scores, aes(x = PC1, y = PC3, group = pop, colour = pop), type = "norm", linewidth = 0.4, alpha = 0.15, show.legend = FALSE) +
        geom_point(size = 2.1, alpha = 0.9) +
        scale_colour_manual(values = population_colours, drop = FALSE) +
        labs(
                title = "PC1 vs PC3",
                x = paste0("PC1 (", ddrad_explained_variance$explained_variance_percent[1], "%)"),
                y = paste0("PC3 (", ddrad_explained_variance$explained_variance_percent[3], "%)"),
                colour = "Population",
                shape = "Elevation"
        )

ddrad_pca_plot_pc24 <- ggplot(ddrad_scores, aes(x = PC2, y = PC4, colour = pop, shape = elevation_band)) +
        stat_ellipse(data = ddrad_ellipse_scores, aes(x = PC2, y = PC4, group = pop, colour = pop), type = "norm", linewidth = 0.4, alpha = 0.15, show.legend = FALSE) +
        geom_point(size = 2.1, alpha = 0.9) +
        scale_colour_manual(values = population_colours, drop = FALSE) +
        labs(
                title = "PC2 vs PC4",
                x = paste0("PC2 (", ddrad_explained_variance$explained_variance_percent[2], "%)"),
                y = paste0("PC4 (", ddrad_explained_variance$explained_variance_percent[4], "%)"),
                colour = "Population",
                shape = "Elevation"
        )

ddrad_pca_plot_pc34 <- ggplot(ddrad_scores, aes(x = PC3, y = PC4, colour = pop, shape = elevation_band)) +
        stat_ellipse(data = ddrad_ellipse_scores, aes(x = PC3, y = PC4, group = pop, colour = pop), type = "norm", linewidth = 0.4, alpha = 0.15, show.legend = FALSE) +
        geom_point(size = 2.1, alpha = 0.9) +
        scale_colour_manual(values = population_colours, drop = FALSE) +
        labs(
                title = "PC3 vs PC4",
                x = paste0("PC3 (", ddrad_explained_variance$explained_variance_percent[3], "%)"),
                y = paste0("PC4 (", ddrad_explained_variance$explained_variance_percent[4], "%)"),
                colour = "Population",
                shape = "Elevation"
        )

ddrad_pca_panels <- cowplot::plot_grid(
        ddrad_pca_plot_pc12,
        ddrad_pca_plot_pc13,
        ddrad_pca_plot_pc24,
        ddrad_pca_plot_pc34,
        ncol = 2,
        align = "hv"
)

ddrad_pca_plot <- cowplot::plot_grid(
        cowplot::ggdraw() + cowplot::draw_label("ddRAD population structure", fontface = "bold", x = 0.5, hjust = 0.5),
        ddrad_pca_panels,
        ncol = 1,
        rel_heights = c(0.08, 1)
)

ggsave(file.path(output_dir, "ddrad_pca.png"), ddrad_pca_plot, width = 10, height = 8, dpi = 320)
ddrad_pca_plot
```

## Prepare WGS metadata

Repeat the same metadata preparation steps for the WGS dataset so the two marker systems are handled in parallel.

```{r wgs-metadata}
wgs_genlight <- vcfR2genlight(wgs_vcf)
wgs_sample_ids <- indNames(wgs_genlight)
wgs_populations <- sub("-.*", "", wgs_sample_ids)
wgs_populations[wgs_populations == "s2"] <- "K2S2"

wgs_metadata <- data.frame(
        sample_id = wgs_sample_ids,
        pop = wgs_populations,
        stringsAsFactors = FALSE
)
wgs_metadata$elevation_m <- elevation_table$elevation_m[match(wgs_metadata$pop, elevation_table$pop)]
wgs_metadata$pop <- factor(wgs_metadata$pop, levels = population_order)
wgs_metadata$elevation_band <- ifelse(wgs_metadata$elevation_m >= 1000, "High", "Low")

locNames(wgs_genlight) <- paste(wgs_vcf@fix[, 1], wgs_vcf@fix[, 2], sep = "_")
pop(wgs_genlight) <- wgs_metadata$pop
```

## Run WGS PCA

Calculate the WGS PCA, save the explained variance table, and compute the coordinates needed for later WGS plots.

```{r wgs-pca}
wgs_pca <- glPca(wgs_genlight, nf = 4)

wgs_explained_variance <- data.frame(
        principal_component = seq_along(wgs_pca$eig),
        explained_variance_percent = round(100 * wgs_pca$eig / sum(wgs_pca$eig), 2)
)
wgs_explained_variance$principal_component_label <- paste0("PC", wgs_explained_variance$principal_component)

wgs_scores <- as.data.frame(wgs_pca$scores[, 1:4, drop = FALSE])
wgs_scores$sample_id <- wgs_metadata$sample_id
wgs_scores$pop <- wgs_metadata$pop
wgs_scores$elevation_band <- wgs_metadata$elevation_band

wgs_ellipse_scores <- wgs_scores |>
        dplyr::filter(is.finite(PC1), is.finite(PC2), !is.na(pop)) |>
        dplyr::group_by(pop) |>
        dplyr::filter(dplyr::n() >= 3) |>
        dplyr::ungroup()

write.table(
        wgs_explained_variance,
        file = file.path(output_dir, "wgs_pca_explained_variance.tsv"),
        sep = "\t",
        quote = FALSE,
        row.names = FALSE
)

wgs_explained_variance
```

## Plot WGS eigenvalues

This scree-style plot shows how much variation is explained by each principal component in the WGS dataset.

Question: Compared with ddRAD, does the WGS PCA show a steeper drop after the first few axes or a more gradual decline?

```{r wgs-eigenvalues, fig.width=6.5, fig.height=4.5}
wgs_eigenvalue_data <- as.data.frame(wgs_explained_variance)
wgs_eigenvalue_data$principal_component_num <- suppressWarnings(as.integer(sub("^PC", "", wgs_eigenvalue_data$principal_component)))
if (any(is.na(wgs_eigenvalue_data$principal_component_num))) {
        wgs_eigenvalue_data$principal_component_num <- seq_len(nrow(wgs_eigenvalue_data))
}
wgs_eigenvalue_data$principal_component_label <- paste0("PC", wgs_eigenvalue_data$principal_component_num)

wgs_eigenvalue_plot <- ggplot(wgs_eigenvalue_data, aes(x = principal_component_num, y = explained_variance_percent)) +
        geom_col(fill = "#7570B3", width = 0.75) +
        geom_text(aes(label = sprintf("%.1f", explained_variance_percent)), vjust = -0.3, size = 3) +
        scale_x_continuous(
                breaks = wgs_eigenvalue_data$principal_component_num[seq(1, nrow(wgs_eigenvalue_data), by = 5)],
                labels = wgs_eigenvalue_data$principal_component_label[seq(1, nrow(wgs_eigenvalue_data), by = 5)]
        ) +
        labs(title = "WGS PCA eigenvalues", x = "Principal component", y = "Explained variance (%)") +
        theme_minimal(base_size = 11)

ggsave(file.path(output_dir, "wgs_pca_eigenvalues.png"), wgs_eigenvalue_plot, width = 6.5, height = 4.5, dpi = 320)
wgs_eigenvalue_plot
```

## Plot WGS loadings

This plot shows the full distribution of WGS marker loadings across the first four PCA axes. Broad dense bands near zero indicate many small contributions, while sharp spikes reveal loci with unusually strong influence on a given axis. The exported top-loading table still summarizes the strongest markers on PC1 and PC2.

Question: Compared with ddRAD, does the WGS loading pattern look more diffuse or more dominated by a few strong peaks?

```{r wgs-loadings, fig.width=9, fig.height=6}
wgs_loading_matrix <- as.data.frame(wgs_pca$loadings[, 1:4, drop = FALSE])
colnames(wgs_loading_matrix) <- paste0("PC", seq_len(ncol(wgs_loading_matrix)))
wgs_loading_matrix$marker <- rownames(wgs_loading_matrix)
wgs_loading_matrix$marker_index <- seq_len(nrow(wgs_loading_matrix))

wgs_loading_table <- tidyr::pivot_longer(
        wgs_loading_matrix[, c("marker", "PC1", "PC2")],
        cols = c("PC1", "PC2"),
        names_to = "axis",
        values_to = "loading"
)
wgs_loading_table$abs_loading <- abs(wgs_loading_table$loading)
wgs_loading_table <- wgs_loading_table |>
        dplyr::group_by(axis) |>
        dplyr::slice_max(order_by = abs_loading, n = 20, with_ties = FALSE) |>
        dplyr::ungroup() |>
        dplyr::arrange(axis, dplyr::desc(abs_loading))

wgs_loading_plot_data <- tidyr::pivot_longer(
        wgs_loading_matrix[, c("marker_index", "PC1", "PC2", "PC3", "PC4")],
        cols = c("PC1", "PC2", "PC3", "PC4"),
        names_to = "axis",
        values_to = "loading"
)

write.table(
        wgs_loading_table,
        file = file.path(output_dir, "wgs_pca_top_loadings.tsv"),
        sep = "\t",
        quote = FALSE,
        row.names = FALSE
)

wgs_loading_plot <- ggplot(wgs_loading_plot_data, aes(x = marker_index, y = loading)) +
        geom_bin2d(bins = 70) +
        facet_wrap(~axis, scales = "free_y") +
        scale_fill_viridis_c(option = "C", name = "Count") +
        labs(title = "WGS PCA loading distributions", x = "Marker index", y = "Loading") +
        theme_minimal(base_size = 11)

ggsave(file.path(output_dir, "wgs_pca_top_loadings.png"), wgs_loading_plot, width = 9, height = 6, dpi = 320)
wgs_loading_plot
```

## Plot WGS PCA

These four panels show how individuals cluster in WGS space across several pairs of principal components. This makes it easier to compare whether later axes reveal additional structure beyond the dominant PC1-PC2 pattern.

```{r wgs-pca-plot, fig.width=10, fig.height=8}
wgs_pca_plot_pc12 <- ggplot(wgs_scores, aes(x = PC1, y = PC2, colour = pop, shape = elevation_band)) +
        stat_ellipse(data = wgs_ellipse_scores, aes(x = PC1, y = PC2, group = pop, colour = pop), type = "norm", linewidth = 0.4, alpha = 0.15, show.legend = FALSE) +
        geom_point(size = 2.1, alpha = 0.9) +
        scale_colour_manual(values = population_colours, drop = FALSE) +
        labs(
                title = "PC1 vs PC2",
                x = paste0("PC1 (", wgs_explained_variance$explained_variance_percent[1], "%)"),
                y = paste0("PC2 (", wgs_explained_variance$explained_variance_percent[2], "%)"),
                colour = "Population",
                shape = "Elevation"
        )

wgs_pca_plot_pc13 <- ggplot(wgs_scores, aes(x = PC1, y = PC3, colour = pop, shape = elevation_band)) +
        stat_ellipse(data = wgs_ellipse_scores, aes(x = PC1, y = PC3, group = pop, colour = pop), type = "norm", linewidth = 0.4, alpha = 0.15, show.legend = FALSE) +
        geom_point(size = 2.1, alpha = 0.9) +
        scale_colour_manual(values = population_colours, drop = FALSE) +
        labs(
                title = "PC1 vs PC3",
                x = paste0("PC1 (", wgs_explained_variance$explained_variance_percent[1], "%)"),
                y = paste0("PC3 (", wgs_explained_variance$explained_variance_percent[3], "%)"),
                colour = "Population",
                shape = "Elevation"
        )

wgs_pca_plot_pc24 <- ggplot(wgs_scores, aes(x = PC2, y = PC4, colour = pop, shape = elevation_band)) +
        stat_ellipse(data = wgs_ellipse_scores, aes(x = PC2, y = PC4, group = pop, colour = pop), type = "norm", linewidth = 0.4, alpha = 0.15, show.legend = FALSE) +
        geom_point(size = 2.1, alpha = 0.9) +
        scale_colour_manual(values = population_colours, drop = FALSE) +
        labs(
                title = "PC2 vs PC4",
                x = paste0("PC2 (", wgs_explained_variance$explained_variance_percent[2], "%)"),
                y = paste0("PC4 (", wgs_explained_variance$explained_variance_percent[4], "%)"),
                colour = "Population",
                shape = "Elevation"
        )

wgs_pca_plot_pc34 <- ggplot(wgs_scores, aes(x = PC3, y = PC4, colour = pop, shape = elevation_band)) +
        stat_ellipse(data = wgs_ellipse_scores, aes(x = PC3, y = PC4, group = pop, colour = pop), type = "norm", linewidth = 0.4, alpha = 0.15, show.legend = FALSE) +
        geom_point(size = 2.1, alpha = 0.9) +
        scale_colour_manual(values = population_colours, drop = FALSE) +
        labs(
                title = "PC3 vs PC4",
                x = paste0("PC3 (", wgs_explained_variance$explained_variance_percent[3], "%)"),
                y = paste0("PC4 (", wgs_explained_variance$explained_variance_percent[4], "%)"),
                colour = "Population",
                shape = "Elevation"
        )

wgs_pca_panels <- cowplot::plot_grid(
        wgs_pca_plot_pc12,
        wgs_pca_plot_pc13,
        wgs_pca_plot_pc24,
        wgs_pca_plot_pc34,
        ncol = 2,
        align = "hv"
)

wgs_pca_plot <- cowplot::plot_grid(
        cowplot::ggdraw() + cowplot::draw_label("WGS population structure", fontface = "bold", x = 0.5, hjust = 0.5),
        wgs_pca_panels,
        ncol = 1,
        rel_heights = c(0.08, 1)
)

ggsave(file.path(output_dir, "wgs_pca.png"), wgs_pca_plot, width = 10, height = 8, dpi = 320)
wgs_pca_plot
```

## Compare the PCA panels

Place the ddRAD and WGS PCA panels side by side so students can compare whether the two marker systems recover similar clustering patterns across multiple axis pairs.

```{r pca-panels, fig.width=11, fig.height=5.5}
pca_overview <- cowplot::plot_grid(ddrad_pca_plot, wgs_pca_plot, ncol = 2, labels = c("A", "B"))

ggsave(file.path(output_dir, "pca_overview.png"), pca_overview, width = 11, height = 5.5, dpi = 320)

pca_overview
```

## Calculate pairwise Fst

Convert both datasets to `genind` objects and calculate the pairwise Fst matrix for each marker system.

```{r genind-and-fst}
ddrad_genind <- vcfR2genind(ddrad_vcf)
pop(ddrad_genind) <- ddrad_metadata$pop
ddrad_fst <- as.matrix(genet.dist(ddrad_genind, method = "Fst"))

wgs_genind <- vcfR2genind(wgs_vcf)
pop(wgs_genind) <- wgs_metadata$pop
wgs_fst <- as.matrix(genet.dist(wgs_genind, method = "Fst"))

reference_order <- population_order[population_order %in% rownames(ddrad_fst)]
ddrad_fst <- ddrad_fst[reference_order, reference_order, drop = FALSE]
wgs_fst <- wgs_fst[reference_order, reference_order, drop = FALSE]
geo_matrix <- geo_matrix[reference_order, reference_order, drop = FALSE]

write.table(ddrad_fst, file = file.path(output_dir, "ddrad_pairwise_fst.tsv"), sep = "\t", quote = FALSE, col.names = NA)
write.table(wgs_fst, file = file.path(output_dir, "wgs_pairwise_fst.tsv"), sep = "\t", quote = FALSE, col.names = NA)

ddrad_fst
```

## Prepare Fst heatmap tables

Convert both Fst matrices to long format so they can be plotted as heatmaps.

```{r fst-heatmap-data}
ddrad_fst_long <- as.data.frame(as.table(ddrad_fst), stringsAsFactors = FALSE)
names(ddrad_fst_long) <- c("population_1", "population_2", "value")
ddrad_fst_long$population_1 <- factor(ddrad_fst_long$population_1, levels = reference_order)
ddrad_fst_long$population_2 <- factor(ddrad_fst_long$population_2, levels = rev(reference_order))

wgs_fst_long <- as.data.frame(as.table(wgs_fst), stringsAsFactors = FALSE)
names(wgs_fst_long) <- c("population_1", "population_2", "value")
wgs_fst_long$population_1 <- factor(wgs_fst_long$population_1, levels = reference_order)
wgs_fst_long$population_2 <- factor(wgs_fst_long$population_2, levels = rev(reference_order))
```

## Plot ddRAD Fst heatmap

This heatmap shows the relative genetic differentiation among populations in the ddRAD dataset.

```{r ddrad-fst-heatmap, fig.width=6, fig.height=5}
ddrad_heatmap <- ggplot(ddrad_fst_long, aes(x = population_1, y = population_2, fill = value)) +
        geom_tile(colour = "white", linewidth = 0.35) +
        geom_text(aes(label = sprintf("%.3f", value)), size = 3) +
        scale_fill_gradient(low = "#F7FBFF", high = "#08306B") +
        labs(title = "ddRAD pairwise differentiation", x = NULL, y = NULL, fill = "Fst") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1), panel.grid = element_blank())

ddrad_heatmap
```

## Plot WGS Fst heatmap

This heatmap shows the relative genetic differentiation among populations in the WGS dataset.

```{r wgs-fst-heatmap, fig.width=6, fig.height=5}
wgs_heatmap <- ggplot(wgs_fst_long, aes(x = population_1, y = population_2, fill = value)) +
        geom_tile(colour = "white", linewidth = 0.35) +
        geom_text(aes(label = sprintf("%.3f", value)), size = 3) +
        scale_fill_gradient(low = "#F7FBFF", high = "#08306B") +
        labs(title = "WGS pairwise differentiation", x = NULL, y = NULL, fill = "Fst") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1), panel.grid = element_blank())

wgs_heatmap
```

## Compare the Fst heatmaps

Place the ddRAD and WGS heatmaps side by side to compare the overall differentiation pattern.

```{r fst-heatmap-panel, fig.width=11, fig.height=5.5}
fst_panel <- cowplot::plot_grid(ddrad_heatmap, wgs_heatmap, ncol = 2, labels = c("A", "B"))

ggsave(file.path(output_dir, "pairwise_fst_heatmaps.png"), fst_panel, width = 11, height = 5.5, dpi = 320)
fst_panel
```

## Build the ddRAD versus WGS Fst table

Keep one value per population pair so you can compare ddRAD Fst and WGS Fst directly in a scatterplot.

```{r fst-correlation-table}
pair_index <- which(lower.tri(ddrad_fst), arr.ind = TRUE)
fst_correlation_table <- data.frame(
        population_1 = rownames(ddrad_fst)[pair_index[, 1]],
        population_2 = colnames(ddrad_fst)[pair_index[, 2]],
        ddrad_fst = ddrad_fst[lower.tri(ddrad_fst)],
        wgs_fst = wgs_fst[lower.tri(wgs_fst)],
        stringsAsFactors = FALSE
)

write.table(
        fst_correlation_table,
        file = file.path(output_dir, "ddrad_vs_wgs_fst_pairs.tsv"),
        sep = "\t",
        quote = FALSE,
        row.names = FALSE
)

fst_correlation_table
```

## Plot ddRAD versus WGS Fst

If the two marker systems tell a similar story, the points should follow a positive trend.

```{r fst-correlation-plot, fig.width=7, fig.height=5.5}
fst_correlation_plot <- ggplot(fst_correlation_table, aes(x = ddrad_fst, y = wgs_fst)) +
        geom_point(size = 2.6, alpha = 0.85, colour = "#264653") +
        geom_smooth(method = "lm", se = FALSE, linewidth = 0.5, colour = "black") +
        geom_text(aes(label = paste(population_1, population_2, sep = "-")), nudge_y = 0.002, size = 3, check_overlap = TRUE) +
        labs(title = "ddRAD versus WGS pairwise Fst", x = "ddRAD pairwise Fst", y = "WGS pairwise Fst")

ggsave(file.path(output_dir, "ddrad_vs_wgs_fst_correlation.png"), fst_correlation_plot, width = 7, height = 5.5, dpi = 320)
fst_correlation_plot
```

## Prepare Mantel test matrices

Create the elevation distance matrix and save the geographic and elevation distance tables used in the Mantel tests.

```{r mantel-inputs}
elevation_matrix <- as.matrix(dist(elevation_table$elevation_m, method = "euclidean", diag = TRUE))
rownames(elevation_matrix) <- elevation_table$pop
colnames(elevation_matrix) <- elevation_table$pop
elevation_matrix <- elevation_matrix[reference_order, reference_order, drop = FALSE]

write.table(geo_matrix, file = file.path(output_dir, "geographic_distance_matrix.tsv"), sep = "\t", quote = FALSE, col.names = NA)
write.table(elevation_matrix, file = file.path(output_dir, "elevation_distance_matrix.tsv"), sep = "\t", quote = FALSE, col.names = NA)
```

## Test isolation by distance

Run Mantel tests to ask whether geographically distant populations are also more genetically differentiated.

```{r ibd-tests}
ddrad_ibd <- mantel(geo_matrix, ddrad_fst, permutations = 1000)
wgs_ibd <- mantel(geo_matrix, wgs_fst, permutations = 1000)

data.frame(
        analysis = c("ddRAD IBD", "WGS IBD"),
        statistic = c(unname(ddrad_ibd$statistic), unname(wgs_ibd$statistic)),
        p_value = c(ddrad_ibd$signif, wgs_ibd$signif)
)
```

## Prepare IBD scatterplots

Extract the pairwise values needed to plot geographic distance against Fst for both datasets.

```{r ibd-plot-data}
ddrad_geo_pairs <- data.frame(
        predictor = geo_matrix[upper.tri(geo_matrix)],
        fst = ddrad_fst[upper.tri(ddrad_fst)]
)

wgs_geo_pairs <- data.frame(
        predictor = geo_matrix[upper.tri(geo_matrix)],
        fst = wgs_fst[upper.tri(wgs_fst)]
)
```

```{r ddrad-ibd-plot, fig.width=6.5, fig.height=5}
ddrad_geo_plot <- ggplot(ddrad_geo_pairs, aes(x = predictor, y = fst)) +
        geom_point(size = 2.6, alpha = 0.85, colour = "#1B9E77") +
        geom_smooth(method = "lm", se = FALSE, linewidth = 0.5, colour = "black") +
        annotate("label", x = Inf, y = Inf, hjust = 1.02, vjust = 1.3,
                 label = sprintf("Mantel r = %.3f\np = %.3f", unname(ddrad_ibd$statistic), ddrad_ibd$signif),
                 size = 3, linewidth = 0.15, fill = "white") +
        labs(title = "ddRAD: isolation by distance", x = "Geographic distance (km)", y = "Pairwise Fst")

ddrad_geo_plot
```

```{r wgs-ibd-plot, fig.width=6.5, fig.height=5}
wgs_geo_plot <- ggplot(wgs_geo_pairs, aes(x = predictor, y = fst)) +
        geom_point(size = 2.6, alpha = 0.85, colour = "#7570B3") +
        geom_smooth(method = "lm", se = FALSE, linewidth = 0.5, colour = "black") +
        annotate("label", x = Inf, y = Inf, hjust = 1.02, vjust = 1.3,
                 label = sprintf("Mantel r = %.3f\np = %.3f", unname(wgs_ibd$statistic), wgs_ibd$signif),
                 size = 3, linewidth = 0.15, fill = "white") +
        labs(title = "WGS: isolation by distance", x = "Geographic distance (km)", y = "Pairwise Fst")

wgs_geo_plot
```

```{r ibe-tests}
ddrad_ibe <- mantel.partial(ddrad_fst, elevation_matrix, geo_matrix, permutations = 1000)
wgs_ibe <- mantel.partial(wgs_fst, elevation_matrix, geo_matrix, permutations = 1000)

test_summary <- data.frame(
        analysis = c(
                "ddRAD IBD",
                "WGS IBD",
                "ddRAD IBE controlling for geography",
                "WGS IBE controlling for geography"
        ),
        statistic = c(
                unname(ddrad_ibd$statistic),
                unname(wgs_ibd$statistic),
                unname(ddrad_ibe$statistic),
                unname(wgs_ibe$statistic)
        ),
        p_value = c(ddrad_ibd$signif, wgs_ibd$signif, ddrad_ibe$signif, wgs_ibe$signif)
)

write.table(
        test_summary,
        file = file.path(output_dir, "mantel_test_summary.tsv"),
        sep = "\t",
        quote = FALSE,
        row.names = FALSE
)

test_summary
```

```{r ibe-plot-data}
ddrad_elev_pairs <- data.frame(
        predictor = elevation_matrix[upper.tri(elevation_matrix)],
        fst = ddrad_fst[upper.tri(ddrad_fst)]
)

wgs_elev_pairs <- data.frame(
        predictor = elevation_matrix[upper.tri(elevation_matrix)],
        fst = wgs_fst[upper.tri(wgs_fst)]
)
```

```{r ddrad-ibe-plot, fig.width=6.5, fig.height=5}
ddrad_elev_plot <- ggplot(ddrad_elev_pairs, aes(x = predictor, y = fst)) +
        geom_point(size = 2.6, alpha = 0.85, colour = "#D95F02") +
        geom_smooth(method = "lm", se = FALSE, linewidth = 0.5, colour = "black") +
        annotate("label", x = Inf, y = Inf, hjust = 1.02, vjust = 1.3,
                 label = sprintf("Partial Mantel r = %.3f\np = %.3f", unname(ddrad_ibe$statistic), ddrad_ibe$signif),
                 size = 3, linewidth = 0.15, fill = "white") +
        labs(title = "ddRAD: elevation contrast", x = "Elevation difference (m)", y = "Pairwise Fst")

ddrad_elev_plot
```

```{r wgs-ibe-plot, fig.width=6.5, fig.height=5}
wgs_elev_plot <- ggplot(wgs_elev_pairs, aes(x = predictor, y = fst)) +
        geom_point(size = 2.6, alpha = 0.85, colour = "#E7298A") +
        geom_smooth(method = "lm", se = FALSE, linewidth = 0.5, colour = "black") +
        annotate("label", x = Inf, y = Inf, hjust = 1.02, vjust = 1.3,
                 label = sprintf("Partial Mantel r = %.3f\np = %.3f", unname(wgs_ibe$statistic), wgs_ibe$signif),
                 size = 3, linewidth = 0.15, fill = "white") +
        labs(title = "WGS: elevation contrast", x = "Elevation difference (m)", y = "Pairwise Fst")

wgs_elev_plot
```

```{r mantel-panels, fig.width=11, fig.height=8}
mantel_panel <- cowplot::plot_grid(
        ddrad_geo_plot,
        wgs_geo_plot,
        ddrad_elev_plot,
        wgs_elev_plot,
        ncol = 2,
        labels = c("A", "B", "C", "D")
)

ggsave(file.path(output_dir, "mantel_relationships.png"), mantel_panel, width = 11, height = 8, dpi = 320)
mantel_panel
```

