---
title: "Practical 3 Part 2: ddRAD and WGS comparison"
output: html_document
---

## Hide setup details


This first chunk only sets global notebook options. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
knitr::opts_knit$set(root.dir = "/home/rstudio/work")
```

## Load packages

Load the R packages needed for PCA, diversity summaries, Mantel tests, and plotting.

```{r libraries}
library(adegenet)
library(cowplot)
library(dplyr)
library(ggplot2)
library(hierfstat)
library(tidyr)
library(vcfR)
library(vegan)
```

## Set plot theme and paths

Define the shared plotting style, the output folder, and the population order used throughout the analysis.

```{r options}
theme_set(
        theme_minimal(base_size = 11) +
                theme(
                        plot.title.position = "plot",
                        legend.position = "bottom",
                        panel.grid.minor = element_blank()
                )
)

output_dir <- "/home/rstudio/work/outputs/02-ddrad-wgs-population-comparison"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

population_order <- c("A2G", "A2S", "V2B", "B2MV", "P2", "K2S2")
population_colours <- c(
        A2G = "#1B9E77",
        A2S = "#D95F02",
        V2B = "#7570B3",
        B2MV = "#E7298A",
        P2 = "#66A61E",
        K2S2 = "#E6AB02"
)
```

## Read the input data

Load the elevation table, the geographic distance matrix, and the ddRAD and WGS VCF files.

```{r read-inputs}
elevation_table <- read.table(
        "/home/rstudio/work/dataset-tutorial3/02-ddrad-wgs-population-comparison/elevation.txt",
        header = FALSE,
        stringsAsFactors = FALSE
)
names(elevation_table) <- c("pop", "elevation_m")

geo_table <- read.table(
        "/home/rstudio/work/dataset-tutorial3/02-ddrad-wgs-population-comparison/geo_dist.txt",
        header = TRUE,
        check.names = FALSE
)
geo_matrix <- as.matrix(geo_table)

ddrad_vcf <- read.vcfR(
        "/home/rstudio/work/dataset-tutorial3/02-ddrad-wgs-population-comparison/ddRAD_filtered_pruned.vcf.gz",
        verbose = FALSE
)

wgs_vcf <- read.vcfR(
        "/home/rstudio/work/dataset-tutorial3/02-ddrad-wgs-population-comparison/diploids_bisnps_filt1rem_minDP4nc_maxDP1009rm_filtMAC4MAF01_MD02_pruned.vcf.gz",
        verbose = FALSE
)
```

## Prepare ddRAD metadata

Extract sample names and population labels from the ddRAD VCF, then attach elevation information for plotting.

```{r ddrad-metadata}
ddrad_genlight <- vcfR2genlight(ddrad_vcf)
ddrad_sample_ids <- indNames(ddrad_genlight)
ddrad_populations <- sub("-.*", "", ddrad_sample_ids)
ddrad_populations[ddrad_populations == "s2"] <- "K2S2"

ddrad_metadata <- data.frame(
        sample_id = ddrad_sample_ids,
        pop = ddrad_populations,
        stringsAsFactors = FALSE
)
ddrad_metadata$elevation_m <- elevation_table$elevation_m[match(ddrad_metadata$pop, elevation_table$pop)]
ddrad_metadata$pop <- factor(ddrad_metadata$pop, levels = population_order)
ddrad_metadata$elevation_level <- ifelse(ddrad_metadata$elevation_m >= 1000, "High", "Low")

locNames(ddrad_genlight) <- paste(ddrad_vcf@fix[, 1], ddrad_vcf@fix[, 2], sep = "_")
pop(ddrad_genlight) <- ddrad_metadata$pop
```

## Run ddRAD PCA

Calculate the ddRAD PCA, save the explained variance table, and compute the coordinates needed for later PCA plots.

```{r ddrad-pca}
ddrad_pca <- glPca(ddrad_genlight, nf = 20)

# `ddrad_pca$eig` contains the eigenvalues: one value per PCA axis, where larger values mean that axis captures more variation.
# This table converts those eigenvalues into the percentage of total variation explained by each PCA axis.
ddrad_explained_variance <- data.frame(
        principal_component = seq_along(ddrad_pca$eig),
        explained_variance_percent = round(100 * ddrad_pca$eig / sum(ddrad_pca$eig), 2)
)
ddrad_explained_variance$principal_component_label <- paste0("PC", ddrad_explained_variance$principal_component)

# These scores are the sample coordinates used to place each individual on the PCA plots.
ddrad_scores <- as.data.frame(ddrad_pca$scores[, 1:4, drop = FALSE])
ddrad_scores$sample_id <- ddrad_metadata$sample_id
ddrad_scores$pop <- ddrad_metadata$pop
ddrad_scores$elevation_level <- ddrad_metadata$elevation_level

ddrad_explained_variance
```

## Plot ddRAD eigenvalues

This scree-style plot shows how much variation is explained by each principal component in the ddRAD dataset.

Question: Do the first one or two ddRAD axes capture most of the variation, or is the signal spread across several components?

```{r ddrad-eigenvalues, fig.width=6.5, fig.height=4.5}
ddrad_eigenvalue_data <- as.data.frame(ddrad_explained_variance)

ddrad_eigenvalue_plot <- ggplot(ddrad_eigenvalue_data, aes(x = principal_component, y = explained_variance_percent)) +
        geom_col(fill = "#1B9E77", width = 0.75) +
        geom_text(aes(label = paste(format(round(explained_variance_percent, 1), nsmall = 1))), vjust = -0.3, size = 3) +
        scale_x_continuous(
                breaks = ddrad_eigenvalue_data$principal_component[seq(1, nrow(ddrad_eigenvalue_data), by = 5)],
                labels = ddrad_eigenvalue_data$principal_component_label[seq(1, nrow(ddrad_eigenvalue_data), by = 5)]
        ) +
        labs(title = "ddRAD PCA eigenvalues", x = "Principal component", y = "Explained variance (%)") +
        theme_minimal(base_size = 11)

ddrad_eigenvalue_plot
```

## Plot ddRAD loadings

This plot shows the full distribution of ddRAD marker loadings across the first four PCA axes. Dense horizontal bands near zero mean that most markers contribute weakly, whereas isolated vertical spikes indicate a smaller number of especially influential markers. The exported top-loading table lists the 20 markers with the strongest absolute contributions on the first two axes.

Question: Do the first ddRAD axes look driven by a few concentrated loading peaks, or by many markers with modest effects?

```{r ddrad-loadings, fig.width=9, fig.height=6}
ddrad_loading_matrix <- as.data.frame(ddrad_pca$loadings[, 1:4, drop = FALSE])
colnames(ddrad_loading_matrix) <- paste0("PC", seq_len(ncol(ddrad_loading_matrix)))
ddrad_loading_matrix$marker <- rownames(ddrad_loading_matrix)
ddrad_loading_matrix$marker_index <- seq_len(nrow(ddrad_loading_matrix))

ddrad_loading_table <- tidyr::pivot_longer(
        ddrad_loading_matrix[, c("marker", "PC1", "PC2")],
        cols = c("PC1", "PC2"),
        names_to = "axis",
        values_to = "loading"
)
ddrad_loading_table$abs_loading <- abs(ddrad_loading_table$loading)
ddrad_loading_table <- ddrad_loading_table %>%
        dplyr::arrange(axis, dplyr::desc(abs_loading)) %>%
        dplyr::group_by(axis) %>%
        dplyr::slice_head(n = 20) %>%
        dplyr::ungroup()

ddrad_loading_plot_data <- data.frame(
        marker_index = rep(ddrad_loading_matrix$marker_index, times = 4),
        axis = rep(c("PC1", "PC2", "PC3", "PC4"), each = nrow(ddrad_loading_matrix)),
        loading = c(ddrad_loading_matrix$PC1, ddrad_loading_matrix$PC2, ddrad_loading_matrix$PC3, ddrad_loading_matrix$PC4)
)

ddrad_loading_plot <- ggplot(ddrad_loading_plot_data, aes(x = marker_index, y = loading)) +
        geom_bin2d(bins = 70) +
        facet_wrap(~axis, scales = "free_y") +
        scale_fill_viridis_c(option = "C", name = "Count") +
        labs(title = "ddRAD PCA loading distributions", x = "Marker index", y = "Loading") +
        theme_minimal(base_size = 11)
ddrad_loading_plot
```

## Plot ddRAD PC1 and PC2

Start with plotting the first two axes, because this is usually the easiest place to see the dominant clustering pattern.

```{r ddrad-pca-plot-pc12, fig.width=6.5, fig.height=5}
ddrad_pca_plot_pc12 <- ggplot(ddrad_scores, aes(x = PC1, y = PC2, colour = pop, shape = elevation_level)) +
        geom_point(size = 2.1, alpha = 0.9) +
        scale_colour_manual(values = population_colours, drop = FALSE) +
        labs(
                title = "PC1 vs PC2",
                x = paste0("PC1 (", ddrad_explained_variance$explained_variance_percent[1], "%)"),
                y = paste0("PC2 (", ddrad_explained_variance$explained_variance_percent[2], "%)"),
                colour = "Population",
                shape = "Elevation"
        )

ggsave(file.path(output_dir, "ddrad_pca.png"), ddrad_pca_plot_pc12, width = 6.5, height = 5, dpi = 320)
ddrad_pca_plot_pc12
```

## Plot additional ddRAD axes

If the first ddRAD panel does not capture the full pattern, look at the next axis combinations to see whether additional structure appears on later components.

```{r ddrad-pca-other-plots, fig.width=10, fig.height=6.5}
ddrad_pca_plot_pc13 <- ggplot(ddrad_scores, aes(x = PC1, y = PC3, colour = pop, shape = elevation_level)) +
        geom_point(size = 2.1, alpha = 0.9) +
        scale_colour_manual(values = population_colours, drop = FALSE) +
        labs(
                title = "PC1 vs PC3",
                x = paste0("PC1 (", ddrad_explained_variance$explained_variance_percent[1], "%)"),
                y = paste0("PC3 (", ddrad_explained_variance$explained_variance_percent[3], "%)"),
                colour = "Population",
                shape = "Elevation"
        )

ddrad_pca_plot_pc24 <- ggplot(ddrad_scores, aes(x = PC2, y = PC4, colour = pop, shape = elevation_level)) +
        geom_point(size = 2.1, alpha = 0.9) +
        scale_colour_manual(values = population_colours, drop = FALSE) +
        labs(
                title = "PC2 vs PC4",
                x = paste0("PC2 (", ddrad_explained_variance$explained_variance_percent[2], "%)"),
                y = paste0("PC4 (", ddrad_explained_variance$explained_variance_percent[4], "%)"),
                colour = "Population",
                shape = "Elevation"
        )

ddrad_pca_plot_pc34 <- ggplot(ddrad_scores, aes(x = PC3, y = PC4, colour = pop, shape = elevation_level)) +
        geom_point(size = 2.1, alpha = 0.9) +
        scale_colour_manual(values = population_colours, drop = FALSE) +
        labs(
                title = "PC3 vs PC4",
                x = paste0("PC3 (", ddrad_explained_variance$explained_variance_percent[3], "%)"),
                y = paste0("PC4 (", ddrad_explained_variance$explained_variance_percent[4], "%)"),
                colour = "Population",
                shape = "Elevation"
        )

ddrad_other_pca_plot <- cowplot::plot_grid(
        ddrad_pca_plot_pc12,
        ddrad_pca_plot_pc13,
        ddrad_pca_plot_pc24,
        ddrad_pca_plot_pc34,
        ncol = 2,
        align = "hv"
)

ddrad_other_pca_plot
```

## Prepare WGS metadata

Repeat the same metadata preparation steps for the WGS dataset so the two marker systems are handled in parallel.

```{r wgs-metadata}
wgs_genlight <- vcfR2genlight(wgs_vcf)
wgs_sample_ids <- indNames(wgs_genlight)
wgs_populations <- sub("-.*", "", wgs_sample_ids)
wgs_populations[wgs_populations == "s2"] <- "K2S2"

wgs_metadata <- data.frame(
        sample_id = wgs_sample_ids,
        pop = wgs_populations,
        stringsAsFactors = FALSE
)
wgs_metadata$elevation_m <- elevation_table$elevation_m[match(wgs_metadata$pop, elevation_table$pop)]
wgs_metadata$pop <- factor(wgs_metadata$pop, levels = population_order)
wgs_metadata$elevation_level <- ifelse(wgs_metadata$elevation_m >= 1000, "High", "Low")

locNames(wgs_genlight) <- paste(wgs_vcf@fix[, 1], wgs_vcf@fix[, 2], sep = "_")
pop(wgs_genlight) <- wgs_metadata$pop
```

## Run WGS PCA

Calculate the WGS PCA, save the explained variance table, and compute the coordinates needed for later WGS plots.

```{r wgs-pca}
wgs_pca <- glPca(wgs_genlight, nf = 20)

# `wgs_pca$eig` contains the eigenvalues: one value per PCA axis, where larger values mean that axis captures more variation.
# This table converts those eigenvalues into the percentage of total variation explained by each PCA axis.
wgs_explained_variance <- data.frame(
        principal_component = seq_along(wgs_pca$eig),
        explained_variance_percent = round(100 * wgs_pca$eig / sum(wgs_pca$eig), 2)
)
wgs_explained_variance$principal_component_label <- paste0("PC", wgs_explained_variance$principal_component)

# These scores are the sample coordinates used to place each individual on the PCA plots.
wgs_scores <- as.data.frame(wgs_pca$scores[, 1:4, drop = FALSE])
wgs_scores$sample_id <- wgs_metadata$sample_id
wgs_scores$pop <- wgs_metadata$pop
wgs_scores$elevation_level <- wgs_metadata$elevation_level

wgs_explained_variance
```

## Plot WGS eigenvalues

This scree-style plot shows how much variation is explained by each principal component in the WGS dataset.

Question: Compared with ddRAD, does the WGS PCA show a steeper drop after the first few axes or a more gradual decline?

```{r wgs-eigenvalues, fig.width=6.5, fig.height=4.5}
wgs_eigenvalue_data <- as.data.frame(wgs_explained_variance)

wgs_eigenvalue_plot <- ggplot(wgs_eigenvalue_data, aes(x = principal_component, y = explained_variance_percent)) +
        geom_col(fill = "#7570B3", width = 0.75) +
        geom_text(aes(label = paste(format(round(explained_variance_percent, 1), nsmall = 1))), vjust = -0.3, size = 3) +
        scale_x_continuous(
                breaks = wgs_eigenvalue_data$principal_component[seq(1, nrow(wgs_eigenvalue_data), by = 5)],
                labels = wgs_eigenvalue_data$principal_component_label[seq(1, nrow(wgs_eigenvalue_data), by = 5)]
        ) +
        labs(title = "WGS PCA eigenvalues", x = "Principal component", y = "Explained variance (%)") +
        theme_minimal(base_size = 11)

wgs_eigenvalue_plot
```

## Plot WGS loadings

This plot shows the full distribution of WGS marker loadings across the first four PCA axes. Broad dense bands near zero indicate many small contributions, while sharp spikes reveal loci with unusually strong influence on a given axis. The exported top-loading table still summarizes the strongest markers on PC1 and PC2.

Question: Compared with ddRAD, does the WGS loading pattern look more diffuse or more dominated by a few strong peaks?

```{r wgs-loadings, fig.width=9, fig.height=6}
wgs_loading_matrix <- as.data.frame(wgs_pca$loadings[, 1:4, drop = FALSE])
colnames(wgs_loading_matrix) <- paste0("PC", seq_len(ncol(wgs_loading_matrix)))
wgs_loading_matrix$marker <- rownames(wgs_loading_matrix)
wgs_loading_matrix$marker_index <- seq_len(nrow(wgs_loading_matrix))

wgs_loading_table <- tidyr::pivot_longer(
        wgs_loading_matrix[, c("marker", "PC1", "PC2")],
        cols = c("PC1", "PC2"),
        names_to = "axis",
        values_to = "loading"
)
wgs_loading_table$abs_loading <- abs(wgs_loading_table$loading)
wgs_loading_table <- wgs_loading_table %>%
        dplyr::arrange(axis, dplyr::desc(abs_loading)) %>%
        dplyr::group_by(axis) %>%
        dplyr::slice_head(n = 20) %>%
        dplyr::ungroup()

wgs_loading_plot_data <- data.frame(
        marker_index = rep(wgs_loading_matrix$marker_index, times = 4),
        axis = rep(c("PC1", "PC2", "PC3", "PC4"), each = nrow(wgs_loading_matrix)),
        loading = c(wgs_loading_matrix$PC1, wgs_loading_matrix$PC2, wgs_loading_matrix$PC3, wgs_loading_matrix$PC4)
)

wgs_loading_plot <- ggplot(wgs_loading_plot_data, aes(x = marker_index, y = loading)) +
        geom_bin2d(bins = 70) +
        facet_wrap(~axis, scales = "free_y") +
        scale_fill_viridis_c(option = "C", name = "Count") +
        labs(title = "WGS PCA loading distributions", x = "Marker index", y = "Loading") +
        theme_minimal(base_size = 11)
wgs_loading_plot
```

## Plot WGS PCA

These four panels show how individuals cluster in WGS space across several pairs of principal components, so the full WGS pattern can be inspected in one place.

```{r wgs-pca-plot, fig.width=10, fig.height=8}
wgs_pca_plot_pc12 <- ggplot(wgs_scores, aes(x = PC1, y = PC2, colour = pop, shape = elevation_level)) +
        geom_point(size = 2.1, alpha = 0.9) +
        scale_colour_manual(values = population_colours, drop = FALSE) +
        labs(
                title = "PC1 vs PC2",
                x = paste0("PC1 (", wgs_explained_variance$explained_variance_percent[1], "%)"),
                y = paste0("PC2 (", wgs_explained_variance$explained_variance_percent[2], "%)"),
                colour = "Population",
                shape = "Elevation"
        )

wgs_pca_plot_pc13 <- ggplot(wgs_scores, aes(x = PC1, y = PC3, colour = pop, shape = elevation_level)) +
        geom_point(size = 2.1, alpha = 0.9) +
        scale_colour_manual(values = population_colours, drop = FALSE) +
        labs(
                title = "PC1 vs PC3",
                x = paste0("PC1 (", wgs_explained_variance$explained_variance_percent[1], "%)"),
                y = paste0("PC3 (", wgs_explained_variance$explained_variance_percent[3], "%)"),
                colour = "Population",
                shape = "Elevation"
        )

wgs_pca_plot_pc24 <- ggplot(wgs_scores, aes(x = PC2, y = PC4, colour = pop, shape = elevation_level)) +
        geom_point(size = 2.1, alpha = 0.9) +
        scale_colour_manual(values = population_colours, drop = FALSE) +
        labs(
                title = "PC2 vs PC4",
                x = paste0("PC2 (", wgs_explained_variance$explained_variance_percent[2], "%)"),
                y = paste0("PC4 (", wgs_explained_variance$explained_variance_percent[4], "%)"),
                colour = "Population",
                shape = "Elevation"
        )

wgs_pca_plot_pc34 <- ggplot(wgs_scores, aes(x = PC3, y = PC4, colour = pop, shape = elevation_level)) +
        geom_point(size = 2.1, alpha = 0.9) +
        scale_colour_manual(values = population_colours, drop = FALSE) +
        labs(
                title = "PC3 vs PC4",
                x = paste0("PC3 (", wgs_explained_variance$explained_variance_percent[3], "%)"),
                y = paste0("PC4 (", wgs_explained_variance$explained_variance_percent[4], "%)"),
                colour = "Population",
                shape = "Elevation"
        )

wgs_pca_panels <- cowplot::plot_grid(
        wgs_pca_plot_pc12,
        wgs_pca_plot_pc13,
        wgs_pca_plot_pc24,
        wgs_pca_plot_pc34,
        ncol = 2,
        align = "hv"
)

wgs_pca_plot <- cowplot::plot_grid(
        cowplot::ggdraw() + cowplot::draw_label("WGS population structure", fontface = "bold", x = 0.5, hjust = 0.5),
        wgs_pca_panels,
        ncol = 1,
        rel_heights = c(0.08, 1)
)

ggsave(file.path(output_dir, "wgs_pca.png"), wgs_pca_plot, width = 10, height = 8, dpi = 320)
wgs_pca_plot
```

## Calculate pairwise Fst

Convert both datasets to `genind` objects and calculate the pairwise Fst matrix for each marker system.

```{r genind-and-fst}
ddrad_genind <- vcfR2genind(ddrad_vcf)
pop(ddrad_genind) <- ddrad_metadata$pop
ddrad_fst <- as.matrix(genet.dist(ddrad_genind, method = "Fst"))

wgs_genind <- vcfR2genind(wgs_vcf)
pop(wgs_genind) <- wgs_metadata$pop
wgs_fst <- as.matrix(genet.dist(wgs_genind, method = "Fst"))

reference_order <- population_order[population_order %in% rownames(ddrad_fst)]
ddrad_fst <- ddrad_fst[reference_order, reference_order, drop = FALSE]
wgs_fst <- wgs_fst[reference_order, reference_order, drop = FALSE]
geo_matrix <- geo_matrix[reference_order, reference_order, drop = FALSE]

ddrad_fst
wgs_fst
```

## Prepare Fst heatmap tables

Convert both Fst matrices to long format so they can be plotted as heatmaps.

```{r fst-heatmap-data}
ddrad_fst_long <- as.data.frame(as.table(ddrad_fst), stringsAsFactors = FALSE)
names(ddrad_fst_long) <- c("population_1", "population_2", "value")
ddrad_fst_long$population_1 <- factor(ddrad_fst_long$population_1, levels = reference_order)
ddrad_fst_long$population_2 <- factor(ddrad_fst_long$population_2, levels = rev(reference_order))

wgs_fst_long <- as.data.frame(as.table(wgs_fst), stringsAsFactors = FALSE)
names(wgs_fst_long) <- c("population_1", "population_2", "value")
wgs_fst_long$population_1 <- factor(wgs_fst_long$population_1, levels = reference_order)
wgs_fst_long$population_2 <- factor(wgs_fst_long$population_2, levels = rev(reference_order))
```

## Plot Fst heatmap

Plot the relative genetic differentiation among populations in the ddRAD dataset and WGS dataset as heatmaps, using the same colour scale for direct comparison.

```{r fst-heatmap, fig.width=11, fig.height=5.5}
fst_limits <- range(c(ddrad_fst_long$value, wgs_fst_long$value), na.rm = TRUE)
ddrad_heatmap <- ggplot(ddrad_fst_long, aes(x = population_1, y = population_2, fill = value)) +
        geom_tile(colour = "white", linewidth = 0.35) +
        geom_text(aes(label = paste(format(round(value, 3), nsmall = 3))), size = 3) +
        scale_fill_gradient(low = "#F7FBFF", high = "#08306B", limits = fst_limits) +
        labs(title = "ddRAD pairwise differentiation", x = NULL, y = NULL, fill = "Fst") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1), panel.grid = element_blank())


wgs_heatmap <- ggplot(wgs_fst_long, aes(x = population_1, y = population_2, fill = value)) +
        geom_tile(colour = "white", linewidth = 0.35) +
        geom_text(aes(label = paste(format(round(value, 3), nsmall = 3))), size = 3) +
        scale_fill_gradient(low = "#F7FBFF", high = "#08306B", limits = fst_limits) +
        labs(title = "WGS pairwise differentiation", x = NULL, y = NULL, fill = "Fst") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1), panel.grid = element_blank())

fst_panel <- cowplot::plot_grid(ddrad_heatmap, wgs_heatmap, ncol = 2, labels = c("A", "B"))

ggsave(file.path(output_dir, "pairwise_fst_heatmaps.png"), fst_panel, width = 11, height = 5.5, dpi = 320)
fst_panel
```

##  Compare ddRAD and WGS pairwise Fst

Here we ask whether population pairs that are highly differentiated in WGS are also highly differentiated in ddRAD. does ddRAD captures the broad population differentiation pattern as well as WGS.


```{r fst-correlation-table}
pair_index <- which(lower.tri(ddrad_fst), arr.ind = TRUE)
fst_correlation_table <- data.frame(
        population_1 = rownames(ddrad_fst)[pair_index[, 1]],
        population_2 = colnames(ddrad_fst)[pair_index[, 2]],
        ddrad_fst = ddrad_fst[lower.tri(ddrad_fst)],
        wgs_fst = wgs_fst[lower.tri(wgs_fst)],
        stringsAsFactors = FALSE
)

fst_correlation_test <- cor.test(fst_correlation_table$ddrad_fst, fst_correlation_table$wgs_fst)

fst_correlation_plot <- ggplot(fst_correlation_table, aes(x = ddrad_fst, y = wgs_fst)) +
        geom_point(size = 2.6, alpha = 0.85, colour = "#264653") +
        geom_smooth(method = "lm", se = FALSE, linewidth = 0.5, colour = "black") +
        geom_text(aes(label = paste(population_1, population_2, sep = "-")), nudge_y = 0.002, size = 3, check_overlap = TRUE) +
        labs(
                title = "ddRAD versus WGS pairwise Fst",
                subtitle = paste(
                        "Correlation r =", format(round(as.numeric(fst_correlation_test$estimate), 3), nsmall = 3),
                        "| p =", format(round(fst_correlation_test$p.value, 3), nsmall = 3)
                ),
                x = "ddRAD pairwise Fst",
                y = "WGS pairwise Fst"
        )

ggsave(file.path(output_dir, "ddrad_vs_wgs_fst_correlation.png"), fst_correlation_plot, width = 7, height = 5.5, dpi = 320)
fst_correlation_plot
```

## Prepare Mantel test matrices

Create the elevation distance matrix and save the geographic and elevation distance tables used in the Mantel tests.

```{r mantel-inputs}
elevation_matrix <- as.matrix(dist(elevation_table$elevation_m, method = "euclidean", diag = TRUE))
rownames(elevation_matrix) <- elevation_table$pop
colnames(elevation_matrix) <- elevation_table$pop
elevation_matrix <- elevation_matrix[reference_order, reference_order, drop = FALSE]

geo_matrix
elevation_matrix
```

## Test isolation by distance

A Mantel test compares two distance matrices. Here we test whether populations that are geographically farther apart also have higher genetic differentiation.

```{r ibd-tests}
ddrad_ibd <- mantel(geo_matrix, ddrad_fst, permutations = 1000)
wgs_ibd <- mantel(geo_matrix, wgs_fst, permutations = 1000)

data.frame(
        analysis = c("ddRAD IBD", "WGS IBD"),
        statistic = c(as.numeric(ddrad_ibd$statistic), as.numeric(wgs_ibd$statistic)),
        p_value = c(ddrad_ibd$signif, wgs_ibd$signif)
)
```

## Prepare IBD scatterplots

Extract the pairwise values needed to plot geographic distance against Fst for both datasets.

```{r ibd-plot-data}
ddrad_geo_pairs <- data.frame(
        geographic_distance_km = geo_matrix[upper.tri(geo_matrix)],
        fst = ddrad_fst[upper.tri(ddrad_fst)]
)

wgs_geo_pairs <- data.frame(
        geographic_distance_km = geo_matrix[upper.tri(geo_matrix)],
        fst = wgs_fst[upper.tri(wgs_fst)]
)
```

```{r ddrad-ibd-plot, fig.width=6.5, fig.height=5}
ddrad_geo_plot <- ggplot(ddrad_geo_pairs, aes(x = geographic_distance_km, y = fst)) +
        geom_point(size = 2.6, alpha = 0.85, colour = "#1B9E77") +
        geom_smooth(method = "lm", se = FALSE, linewidth = 0.5, colour = "black") +
        annotate("label", x = Inf, y = Inf, hjust = 1.02, vjust = 1.3,
                 label = paste(
                         paste("Mantel r =", format(round(as.numeric(ddrad_ibd$statistic), 3), nsmall = 3)),
                         paste("p =", format(round(ddrad_ibd$signif, 3), nsmall = 3)),
                         sep = "\n"
                 ),
                 size = 3, linewidth = 0.15, fill = "white") +
        labs(title = "ddRAD: isolation by distance", x = "Geographic distance (km)", y = "Pairwise Fst")

ddrad_geo_plot
```

```{r wgs-ibd-plot, fig.width=6.5, fig.height=5}
wgs_geo_plot <- ggplot(wgs_geo_pairs, aes(x = geographic_distance_km, y = fst)) +
        geom_point(size = 2.6, alpha = 0.85, colour = "#7570B3") +
        geom_smooth(method = "lm", se = FALSE, linewidth = 0.5, colour = "black") +
        annotate("label", x = Inf, y = Inf, hjust = 1.02, vjust = 1.3,
                 label = paste(
                         paste("Mantel r =", format(round(as.numeric(wgs_ibd$statistic), 3), nsmall = 3)),
                         paste("p =", format(round(wgs_ibd$signif, 3), nsmall = 3)),
                         sep = "\n"
                 ),
                 size = 3, linewidth = 0.15, fill = "white") +
        labs(title = "WGS: isolation by distance", x = "Geographic distance (km)", y = "Pairwise Fst")

wgs_geo_plot
```

## Isolation by elevation after accounting for geography

This test asks whether elevation differences are associated with genetic differentiation after accounting for geographic distance. 

```{r ibe-tests}
ddrad_ibe <- mantel.partial(ddrad_fst, elevation_matrix, geo_matrix, permutations = 1000)
wgs_ibe <- mantel.partial(wgs_fst, elevation_matrix, geo_matrix, permutations = 1000)

test_summary <- data.frame(
        analysis = c(
                "ddRAD IBD",
                "WGS IBD",
                "ddRAD IBE controlling for geography",
                "WGS IBE controlling for geography"
        ),
        statistic = c(
                as.numeric(ddrad_ibd$statistic),
                as.numeric(wgs_ibd$statistic),
                as.numeric(ddrad_ibe$statistic),
                as.numeric(wgs_ibe$statistic)
        ),
        p_value = c(ddrad_ibd$signif, wgs_ibd$signif, ddrad_ibe$signif, wgs_ibe$signif)
)

write.table(
        test_summary,
        file = file.path(output_dir, "mantel_test_summary.tsv"),
        sep = "\t",
        quote = FALSE,
        row.names = FALSE
)

test_summary
```

```{r ibe-plot-data}
ddrad_elev_pairs <- data.frame(
        elevation_difference_m = elevation_matrix[upper.tri(elevation_matrix)],
        fst = ddrad_fst[upper.tri(ddrad_fst)]
)

wgs_elev_pairs <- data.frame(
        elevation_difference_m = elevation_matrix[upper.tri(elevation_matrix)],
        fst = wgs_fst[upper.tri(wgs_fst)]
)
```

```{r ddrad-ibe-plot, fig.width=6.5, fig.height=5}
ddrad_elev_plot <- ggplot(ddrad_elev_pairs, aes(x = elevation_difference_m, y = fst)) +
        geom_point(size = 2.6, alpha = 0.85, colour = "#D95F02") +
        geom_smooth(method = "lm", se = FALSE, linewidth = 0.5, colour = "black") +
        annotate("label", x = Inf, y = Inf, hjust = 1.02, vjust = 1.3,
                 label = paste(
                         paste("Partial Mantel r =", format(round(as.numeric(ddrad_ibe$statistic), 3), nsmall = 3)),
                         paste("p =", format(round(ddrad_ibe$signif, 3), nsmall = 3)),
                         sep = "\n"
                 ),
                 size = 3, linewidth = 0.15, fill = "white") +
        labs(title = "ddRAD: elevation contrast", x = "Elevation difference (m)", y = "Pairwise Fst")

ddrad_elev_plot
```

```{r wgs-ibe-plot, fig.width=6.5, fig.height=5}
wgs_elev_plot <- ggplot(wgs_elev_pairs, aes(x = elevation_difference_m, y = fst)) +
        geom_point(size = 2.6, alpha = 0.85, colour = "#E7298A") +
        geom_smooth(method = "lm", se = FALSE, linewidth = 0.5, colour = "black") +
        annotate("label", x = Inf, y = Inf, hjust = 1.02, vjust = 1.3,
                 label = paste(
                         paste("Partial Mantel r =", format(round(as.numeric(wgs_ibe$statistic), 3), nsmall = 3)),
                         paste("p =", format(round(wgs_ibe$signif, 3), nsmall = 3)),
                         sep = "\n"
                 ),
                 size = 3, linewidth = 0.15, fill = "white") +
        labs(title = "WGS: elevation contrast", x = "Elevation difference (m)", y = "Pairwise Fst")

wgs_elev_plot
```

```{r mantel-panels, fig.width=11, fig.height=8}
mantel_panel <- cowplot::plot_grid(
        ddrad_geo_plot,
        wgs_geo_plot,
        ddrad_elev_plot,
        wgs_elev_plot,
        ncol = 2,
        labels = c("A", "B", "C", "D")
)

ggsave(file.path(output_dir, "mantel_relationships.png"), mantel_panel, width = 11, height = 8, dpi = 320)
mantel_panel
```

