---
title: "Practical 3 Part 3: scaffold 1 window scan"
output: html_document
---

## Aim

In this notebook we do one simple task twice: calculate window-based $F_{ST}$ along scaffold 1 and compare the pattern between the WGS and ddRAD datasets.

Interpretation guide:

- Higher $F_{ST}$ means stronger genetic differentiation between the two populations in that window.
- FST peaks are candidate regions of elevated relative differentiation, but they can also arise from low within-population diversity, reduced recombination, background selection, local variation in SNP density, or demographic history.
- ddRAD and WGS will not match perfectly because ddRAD samples only a subset of sites, but broad peaks can still be compared.

## Setup

This chunk sets the working directory for the practical and turns off code display.

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
knitr::opts_knit$set(root.dir = "/home/rstudio/work")
```

## Load packages

We only need three packages: one for population-genetic calculations, one for plotting, and one for reading VCF files.

```{r libraries}
library("PopGenome")
library("ggplot2")
library("vcfR")
```

## Choose input files and window settings

We will scan scaffold 1 in 10 kb windows with a 1 kb step.

```{r inputs}
wgs_vcf_file <- "/home/rstudio/work/dataset-tutorial3/03-scaffold1-window-scan/wgs_vcf/scaffold1/A2_wgs_scaffold1.vcf"
wgs_vcf_dir <- "/home/rstudio/work/dataset-tutorial3/03-scaffold1-window-scan/wgs_vcf/scaffold1"

ddrad_vcf_file <- "/home/rstudio/work/dataset-tutorial3/03-scaffold1-window-scan/ddRAD_vcf/scaffold_1/A2_ddRAD_scaffold_1.vcf"
ddrad_vcf_dir <- "/home/rstudio/work/dataset-tutorial3/03-scaffold1-window-scan/ddRAD_vcf/scaffold_1"

window_width <- 10000
window_jump <- 1000
```

## Create output directory
```{r output-dir}
output_dir <- "/home/rstudio/work/outputs/03-scaffold1-window-scan"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

## Helper functions

These helpers keep the ddRAD and WGS code paths consistent when we calculate and plot multiple summary statistics per window.

```{r helpers}
compute_window_stats <- function(genome_slide, dataset_label, variant_positions) {
  genome_slide <- F_ST.stats(genome_slide, mode = "nucleotide")
  fst <- as.data.frame(get.F_ST(genome_slide, mode = FALSE, pairwise = TRUE)[1])
  names(fst) <- "fst"
  fst$fst[fst$fst < 0] <- 0

  fst$start_pos <- as.numeric(sub(" -.*", "", genome_slide@region.names))
  fst$end_pos <- as.numeric(sub(" :", "", sub(".*- ", "", genome_slide@region.names)))
  fst$midpoint_mb <- (fst$start_pos + fst$end_pos) / 2e6
  fst$n_snps <- mapply(
    function(start, end) sum(variant_positions >= start & variant_positions <= end),
    fst$start_pos,
    fst$end_pos
  )

  genome_slide <- diversity.stats(genome_slide, pi = TRUE)
  pi_values <- as.data.frame(genome_slide@Pi)
  names(pi_values) <- c("pi_A2G", "pi_A2S")

  genome_slide <- diversity.stats.between(genome_slide)
  dxy_values <- as.data.frame(genome_slide@nuc.diversity.between)
  names(dxy_values) <- "dxy"

  stats_df <- cbind(fst, pi_values, dxy_values)
  stats_df$dataset <- dataset_label
  stats_df[, c("dataset", "start_pos", "end_pos", "midpoint_mb", "n_snps", "fst", "pi_A2G", "pi_A2S", "dxy")]
}

make_combined_stats_long <- function(stats_df) {
  rbind(
    data.frame(
      dataset = stats_df$dataset,
      midpoint_mb = stats_df$midpoint_mb,
      statistic = "Fst",
      value = stats_df$fst
    ),
    data.frame(
      dataset = stats_df$dataset,
      midpoint_mb = stats_df$midpoint_mb,
      statistic = "pi A2G",
      value = stats_df$pi_A2G
    ),
    data.frame(
      dataset = stats_df$dataset,
      midpoint_mb = stats_df$midpoint_mb,
      statistic = "pi A2S",
      value = stats_df$pi_A2S
    ),
    data.frame(
      dataset = stats_df$dataset,
      midpoint_mb = stats_df$midpoint_mb,
      statistic = "Dxy",
      value = stats_df$dxy
    )
  )
}

plot_combined_stats <- function(stats_df, title_text) {
  stats_long <- make_combined_stats_long(stats_df)

  ggplot(stats_long, aes(x = midpoint_mb, y = value, color = statistic)) +
    geom_smooth(method = "loess", se = FALSE, span = 0.15, linewidth = 1, na.rm = TRUE) +
    labs(
      title = title_text,
      x = "Scaffold 1 position (Mb)",
      y = "Window statistic value",
      color = NULL
    ) +
    scale_color_manual(
      values = c(
        "Fst" = "#d95f02",
        "pi A2G" = "#1b9e77",
        "pi A2S" = "#66a61e",
        "Dxy" = "#7570b3"
      )
    ) +
    theme_classic(base_size = 12)
}
```

## ddRAD step 1: read the VCF and identify samples

We start with ddRAD by reading the VCF and grouping individuals into A2G and A2S.

```{r ddrad-read-vcf}
ddrad_vcf <- read.vcfR(ddrad_vcf_file, verbose = FALSE)
ddrad_genlight <- vcfR2genlight(ddrad_vcf)
ddrad_genlight$ind.names

A2G_ddrad <- grep("A2G", ddrad_genlight$ind.names, value = TRUE)
A2S_ddrad <- grep("A2S", ddrad_genlight$ind.names, value = TRUE)

populations_ddrad <- list(A2G_ddrad, A2S_ddrad)
populations_ddrad
```

## ddRAD step 2: create the PopGenome object

Read the ddRAD directory and assign the two populations.

```{r ddrad-genome-object}
ddrad_genome <- readData(ddrad_vcf_dir, format = "VCF", include.unknown = TRUE)
ddrad_genome <- set.populations(ddrad_genome, populations_ddrad)

ddrad_genome
```

## ddRAD step 3: create sliding windows

Use the same window size and jump so the ddRAD scan can be compared directly to the WGS scan.

```{r ddrad-windows}
ddrad_genome_slide <- sliding.window.transform(
  ddrad_genome,
  width = window_width,
  jump = window_jump,
  type = 2
)
ddrad_genome_slide <- set.populations(ddrad_genome_slide, populations_ddrad, diploid = TRUE)

length(ddrad_genome_slide@region.names)
head(ddrad_genome_slide@region.names)
```

## ddRAD step 4: calculate Fst for each window

Again, we calculate one $F_{ST}$ value per window and set negative values to 0.

```{r ddrad-fst-table}
ddrad_positions <- as.numeric(ddrad_vcf@fix[, "POS"])
ddrad_stats <- compute_window_stats(ddrad_genome_slide, "ddRAD", ddrad_positions)
ddrad_fst <- ddrad_stats[, c("dataset", "start_pos", "end_pos", "midpoint_mb", "n_snps", "fst")]

write.table(
  ddrad_fst,
  file.path(output_dir, "ddrad_scaffold1_window_fst.tsv"),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)

head(ddrad_fst)
```

## Plot the ddRAD scan

This plot is interpreted in the same way as the WGS scan, but ddRAD often looks noisier because fewer markers are available.

```{r ddrad-plot, fig.width=9, fig.height=4}
ddrad_plot <- ggplot(ddrad_fst, aes(x = midpoint_mb, y = fst)) +
  geom_point(color = "#01665e", alpha = 0.55, size = 1.2) +
  geom_smooth(method = "loess", se = FALSE, span = 0.15, color = "#35978f", linewidth = 1) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    title = "ddRAD: scaffold 1 window-based Fst",
    x = "Scaffold 1 position (Mb)",
    y = "Fst between A2G and A2S"
  ) +
  theme_classic(base_size = 12)

ggsave(file.path(output_dir, "ddrad_scaffold1_fst.png"), ddrad_plot, width = 9, height = 4, dpi = 320)

ddrad_plot
```

ddRAD interpretation:

- Expect more irregular point-to-point variation than in WGS.
- If a major peak appears in both datasets, that region is a stronger candidate for real differentiation.
- If a peak appears only in ddRAD, check whether sparse marker coverage could explain it.

## WGS step 1: read the VCF and identify samples

Now do the same step-by-step workflow for the WGS dataset.

```{r wgs-read-vcf}
wgs_vcf <- read.vcfR(wgs_vcf_file, verbose = FALSE)
wgs_genlight <- vcfR2genlight(wgs_vcf)
wgs_genlight$ind.names

A2G_wgs <- grep("A2G", wgs_genlight$ind.names, value = TRUE)
A2S_wgs <- grep("A2S", wgs_genlight$ind.names, value = TRUE)

populations_wgs <- list(A2G_wgs, A2S_wgs)
populations_wgs
```

## WGS step 2: create the PopGenome object

Read the WGS VCF directory into a PopGenome object and assign the two populations.

```{r wgs-genome-object}
wgs_genome <- readData(wgs_vcf_dir, format = "VCF", include.unknown = TRUE)
wgs_genome <- set.populations(wgs_genome, populations_wgs)

wgs_genome
```

## WGS step 3: create sliding windows

Here we split scaffold 1 into overlapping windows of 10 kb, moving by 1 kb each time.

```{r wgs-windows}
wgs_genome_slide <- sliding.window.transform(
  wgs_genome,
  width = window_width,
  jump = window_jump,
  type = 2
)
wgs_genome_slide <- set.populations(wgs_genome_slide, populations_wgs, diploid = TRUE)

length(wgs_genome_slide@region.names)
head(wgs_genome_slide@region.names)
```

## WGS step 4: calculate Fst for each window

This gives one $F_{ST}$ value per window. Negative values are set to 0 because they are usually treated as no differentiation.

```{r wgs-fst-table}
wgs_positions <- as.numeric(wgs_vcf@fix[, "POS"])
wgs_stats <- compute_window_stats(wgs_genome_slide, "WGS", wgs_positions)
wgs_fst <- wgs_stats[, c("dataset", "start_pos", "end_pos", "midpoint_mb", "n_snps", "fst")]

write.table(
  wgs_fst,
  file.path(output_dir, "wgs_scaffold1_window_fst.tsv"),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)

head(wgs_fst)
```

## Plot the WGS scan

The points show individual windows and the smooth line shows the broad trend across the scaffold.

```{r wgs-plot, fig.width=9, fig.height=4}
wgs_plot <- ggplot(wgs_fst, aes(x = midpoint_mb, y = fst)) +
  geom_point(color = "#8c510a", alpha = 0.55, size = 1.2) +
  geom_smooth(method = "loess", se = FALSE, span = 0.15, color = "#bf812d", linewidth = 1) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    title = "WGS: scaffold 1 window-based Fst",
    x = "Scaffold 1 position (Mb)",
    y = "Fst between A2G and A2S"
  ) +
  theme_classic(base_size = 12)

ggsave(file.path(output_dir, "wgs_scaffold1_fst.png"), wgs_plot, width = 9, height = 4, dpi = 320)

wgs_plot
```

WGS interpretation:

- Look for clusters of high windows rather than a single isolated point.
- Regions close to $F_{ST} = 0$ show little differentiation between A2G and A2S.

## Add pi and Dxy to the window scan plots

We can use the same window objects to overlay within-population diversity ($pi$) and between-population divergence (Dxy) on the same scaffold position axis.

```{r wgs-ddrad-combined-stats-tables}
head(ddrad_stats)
head(wgs_stats)
```

These curves are drawn on their raw scales, like in your example script, so the main goal is to compare where peaks and troughs occur along the scaffold.

```{r combined-ddrad-plot, fig.width=9, fig.height=4}
plot_combined_stats(ddrad_stats, "ddRAD: scaffold 1 with Fst, pi, and Dxy")
```

```{r combined-wgs-plot, fig.width=9, fig.height=4}
plot_combined_stats(wgs_stats, "WGS: scaffold 1 with Fst, pi, and Dxy")
```

## What else can PopGenome calculate?

We focus on $F_{ST}$ in this notebook, but PopGenome can calculate many other summary statistics from the same window objects. This chunk is here as a quick guide so you can see what else is possible.

```{r popgenome-help, eval=FALSE}
# Starting point: a windowed PopGenome object such as
# ddrad_genome_slide or wgs_genome_slide

# 1. Within-population nucleotide diversity (pi)
example_pi <- diversity.stats(ddrad_genome_slide, pi = TRUE)
head(as.data.frame(example_pi@Pi))

# 2. Between-population absolute divergence (Dxy)
example_dxy <- diversity.stats.between(example_pi)
head(as.data.frame(example_dxy@nuc.diversity.between))

# 3. Tajima's D and other neutrality statistics
example_tajima <- neutrality.stats(example_dxy, FAST = TRUE)
head(as.data.frame(example_tajima@Tajima.D))

# 4. Haplotype-based statistics
example_haplo <- haplotype.stats(ddrad_genome_slide)

# 5. Linkage disequilibrium
example_ld <- linkage.stats(ddrad_genome_slide)

# 6. Fu and Li statistics
example_fu_li <- neutrality.stats(ddrad_genome_slide, FAST = TRUE)
head(as.data.frame(example_fu_li@Fu.Li.D))
```

Here is one small executed example using the current window objects.

```{r popgenome-example-output}
ddrad_other_stats <- diversity.stats(ddrad_genome_slide, pi = TRUE)
ddrad_other_stats <- diversity.stats.between(ddrad_other_stats)
ddrad_other_stats <- suppressWarnings(neutrality.stats(ddrad_other_stats, FAST = TRUE))

list(
  ddrad_pi = head(as.data.frame(ddrad_other_stats@Pi)),
  ddrad_dxy = head(as.data.frame(ddrad_other_stats@nuc.diversity.between)),
  ddrad_tajima_d = head(as.data.frame(ddrad_other_stats@Tajima.D))
)
```

How to interpret these outputs:

- Higher $pi$ means more genetic diversity within a population.
- Higher Dxy means stronger absolute divergence between populations.
- Tajima's D near 0 is broadly consistent with the standard neutral model under constant population size, but deviations can reflect demography, structure, linked selection, or selection at nearby sites.

## Compare WGS and ddRAD directly

This final chunk keeps the same two-panel comparison, but now each panel overlays $F_{ST}$, $pi$, and Dxy.

```{r compare-plots, fig.width=9, fig.height=7}
combined_all_stats <- rbind(
  make_combined_stats_long(ddrad_stats),
  make_combined_stats_long(wgs_stats)
)

ggplot(combined_all_stats, aes(x = midpoint_mb, y = value, color = statistic)) +
  geom_smooth(method = "loess", se = FALSE, span = 0.15, linewidth = 1, na.rm = TRUE) +
  facet_wrap(~ dataset, ncol = 1) +
  labs(
    title = "Scaffold 1 window scans: ddRAD versus WGS",
    x = "Scaffold 1 position (Mb)",
    y = "Window statistic value",
    color = NULL
  ) +
  scale_color_manual(
    values = c(
      "Fst" = "#d95f02",
      "pi A2G" = "#1b9e77",
      "pi A2S" = "#66a61e",
      "Dxy" = "#7570b3"
    )
  ) +
  theme_classic(base_size = 12)
```

