---
title: "Practical 3 Part 3: scaffold 1 window scan"
output: html_document
---

## Aim

In this notebook we will calculate window-based $F_{ST}$ along scaffold 1 and compare the pattern between the WGS and ddRAD datasets.

Interpretation guide:

- Higher $F_{ST}$ means stronger genetic differentiation between the two populations in that window.
- FST peaks are candidate regions of elevated relative differentiation, but they can also arise from low within-population diversity, reduced recombination, local variation in SNP density, or demographic history.
- ddRAD and WGS will not match perfectly because ddRAD samples only a subset of sites, but broad peaks can still be compared.

## Setup

This chunk sets the working directory for the practical and turns off code display.

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
knitr::opts_knit$set(root.dir = "/home/rstudio/work")
```

## Load packages


```{r libraries}
library("PopGenome")
library("ggplot2")
library("vcfR")
```

## Choose input files and window settings

We will scan scaffold 1 in 10 kb windows with a 1 kb step.

```{r inputs}
wgs_vcf_file <- "/home/rstudio/work/dataset-tutorial3/03-scaffold1-window-scan/wgs_vcf/scaffold1/A2_wgs_scaffold1.vcf"
wgs_vcf_dir <- "/home/rstudio/work/dataset-tutorial3/03-scaffold1-window-scan/wgs_vcf/scaffold1"

ddrad_vcf_file <- "/home/rstudio/work/dataset-tutorial3/03-scaffold1-window-scan/ddRAD_vcf/scaffold_1/A2_ddRAD_scaffold_1.vcf"
ddrad_vcf_dir <- "/home/rstudio/work/dataset-tutorial3/03-scaffold1-window-scan/ddRAD_vcf/scaffold_1"

window_width <- 10000
window_jump <- 1000
```

## Create output directory
```{r output-dir}
output_dir <- "/home/rstudio/work/outputs/03-scaffold1-window-scan"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

plot_colors <- c(
  fst = "#E69F00",
  pi_A2G = "#56B4E9",
  pi_A2S = "#009E73",
  pi_overall = "#CC79A7",
  dxy = "#D55E00"
)

plot_legend_labels <- c(
  fst = "Fst",
  pi_A2G = "pi A2G",
  pi_A2S = "pi A2S",
  pi_overall = "pi overall",
  dxy = "Dxy"
)
```

## Window summary tables

For each dataset we build one table with the window start, window end, midpoint, number of SNPs, $F_{ST}$, $pi$ for each population, and Dxy.


## ddRAD step 1: read the VCF and identify samples

We start with ddRAD by reading the VCF and grouping individuals into A2G and A2S.

```{r ddrad-read-vcf}
ddrad_vcf <- read.vcfR(ddrad_vcf_file, verbose = FALSE)
ddrad_genlight <- vcfR2genlight(ddrad_vcf)
ddrad_genlight$ind.names

A2G_ddrad <- grep("A2G", ddrad_genlight$ind.names, value = TRUE)
A2S_ddrad <- grep("A2S", ddrad_genlight$ind.names, value = TRUE)

populations_ddrad <- list(A2G_ddrad, A2S_ddrad)
populations_ddrad
```

## ddRAD step 2: create the PopGenome object

Read the ddRAD directory and assign the two populations.

```{r ddrad-genome-object}
ddrad_genome <- readData(ddrad_vcf_dir, format = "VCF", include.unknown = TRUE)
ddrad_genome <- set.populations(ddrad_genome, populations_ddrad)

ddrad_genome
```

## ddRAD step 3: create sliding windows

Use the same window size and jump so the ddRAD scan can be compared directly to the WGS scan.

```{r ddrad-windows}
ddrad_genome_slide <- sliding.window.transform(
  ddrad_genome,
  width = window_width,
  jump = window_jump,
  type = 2
)
ddrad_genome_slide <- set.populations(ddrad_genome_slide, populations_ddrad, diploid = TRUE)

length(ddrad_genome_slide@region.names)
head(ddrad_genome_slide@region.names)
```

## ddRAD step 4: calculate Fst, pi, and Dxy for each window

We first calculate one $F_{ST}$ value per window and set negative values to 0. Then we add within-population diversity ($pi$) and between-population divergence (Dxy) to the same table.

```{r ddrad-fst-table}
ddrad_positions <- as.numeric(ddrad_vcf@fix[, "POS"])

ddrad_genome_slide <- F_ST.stats(ddrad_genome_slide, mode = "nucleotide")
ddrad_fst_values <- as.data.frame(get.F_ST(ddrad_genome_slide, mode = FALSE, pairwise = TRUE)[1])
names(ddrad_fst_values) <- "fst"
ddrad_fst_values$fst[ddrad_fst_values$fst < 0] <- 0

ddrad_fst_values$start_pos <- as.numeric(sub(" -.*", "", ddrad_genome_slide@region.names))
ddrad_fst_values$end_pos <- as.numeric(sub(" :", "", sub(".*- ", "", ddrad_genome_slide@region.names)))
ddrad_fst_values$midpoint_mb <- (ddrad_fst_values$start_pos + ddrad_fst_values$end_pos) / 2e6
ddrad_fst_values$n_snps <- mapply(
  function(start, end) sum(ddrad_positions >= start & ddrad_positions <= end),
  ddrad_fst_values$start_pos,
  ddrad_fst_values$end_pos
)

ddrad_genome_slide <- diversity.stats(ddrad_genome_slide, pi = TRUE)
ddrad_pi_values <- as.data.frame(ddrad_genome_slide@Pi)
names(ddrad_pi_values) <- c("pi_A2G", "pi_A2S")

ddrad_genome_slide <- diversity.stats.between(ddrad_genome_slide)
ddrad_dxy_values <- as.data.frame(ddrad_genome_slide@nuc.diversity.between)
names(ddrad_dxy_values) <- "dxy"

ddrad_stats <- cbind(ddrad_fst_values, ddrad_pi_values, ddrad_dxy_values)
ddrad_stats$dataset <- "ddRAD"
ddrad_stats <- ddrad_stats[, c("dataset", "start_pos", "end_pos", "midpoint_mb", "n_snps", "fst", "pi_A2G", "pi_A2S", "dxy")]
ddrad_stats$pi_overall <- rowMeans(ddrad_stats[, c("pi_A2G", "pi_A2S")], na.rm = TRUE)

ddrad_smooth <- list(
  fst = predict(loess(fst ~ midpoint_mb, data = ddrad_stats, span = 0.15, na.action = na.exclude)),
  pi_A2G = predict(loess(pi_A2G ~ midpoint_mb, data = ddrad_stats, span = 0.15, na.action = na.exclude)),
  pi_A2S = predict(loess(pi_A2S ~ midpoint_mb, data = ddrad_stats, span = 0.15, na.action = na.exclude)),
  pi_overall = predict(loess(pi_overall ~ midpoint_mb, data = ddrad_stats, span = 0.15, na.action = na.exclude)),
  dxy = predict(loess(dxy ~ midpoint_mb, data = ddrad_stats, span = 0.15, na.action = na.exclude))
)

ddrad_fst <- ddrad_stats[, c("dataset", "start_pos", "end_pos", "midpoint_mb", "n_snps", "fst")]

write.table(
  ddrad_fst,
  file.path(output_dir, "ddrad_scaffold1_window_fst.tsv"),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)

head(ddrad_fst)
```

## ddRAD step 5: plot the ddRAD window scan

Now plot the ddRAD window summary so you can see where $F_{ST}$, $pi$, and Dxy rise and fall along scaffold 1.

```{r ddrad-plot, fig.width=9, fig.height=6}
ddrad_plot_file <- file.path(output_dir, "ddrad_scaffold1_fst.png")

png(ddrad_plot_file, width = 9, height = 6, units = "in", res = 320)
plot(
  ddrad_stats$midpoint_mb,
  ddrad_stats$fst,
  ylim = c(0, 1),
  main = "Scaffold 1 - ddRAD",
  ylab = "FST",
  xlab = "Position on scaffold 1 (Mb)"
)
lines(ddrad_stats$midpoint_mb, ddrad_smooth$fst, col = plot_colors["fst"], lwd = 3)
lines(ddrad_stats$midpoint_mb, ddrad_smooth$pi_A2G, col = plot_colors["pi_A2G"], lwd = 3)
lines(ddrad_stats$midpoint_mb, ddrad_smooth$pi_A2S, col = plot_colors["pi_A2S"], lwd = 3)
lines(ddrad_stats$midpoint_mb, ddrad_smooth$pi_overall, col = plot_colors["pi_overall"], lwd = 3)
lines(ddrad_stats$midpoint_mb, ddrad_smooth$dxy, col = plot_colors["dxy"], lwd = 3)
legend(
  "topright",
  legend = plot_legend_labels,
  col = plot_colors,
  lwd = 3,
  bty = "n",
  cex = 0.75
)
dev.off()

knitr::include_graphics(ddrad_plot_file)
```

## WGS step 1: read the VCF and identify samples

Now do the same step-by-step workflow for the WGS dataset.

```{r wgs-read-vcf}
wgs_vcf <- read.vcfR(wgs_vcf_file, verbose = FALSE)
wgs_genlight <- vcfR2genlight(wgs_vcf)
wgs_genlight$ind.names

A2G_wgs <- grep("A2G", wgs_genlight$ind.names, value = TRUE)
A2S_wgs <- grep("A2S", wgs_genlight$ind.names, value = TRUE)

populations_wgs <- list(A2G_wgs, A2S_wgs)
populations_wgs
```

## WGS step 2: create the PopGenome object

Read the WGS VCF directory into a PopGenome object and assign the two populations.

```{r wgs-genome-object}
wgs_genome <- readData(wgs_vcf_dir, format = "VCF", include.unknown = TRUE)
wgs_genome <- set.populations(wgs_genome, populations_wgs)

wgs_genome
```

## WGS step 3: create sliding windows

Here we split scaffold 1 into overlapping windows of 10 kb, moving by 1 kb each time.

```{r wgs-windows}
wgs_genome_slide <- sliding.window.transform(
  wgs_genome,
  width = window_width,
  jump = window_jump,
  type = 2
)
wgs_genome_slide <- set.populations(wgs_genome_slide, populations_wgs, diploid = TRUE)

length(wgs_genome_slide@region.names)
head(wgs_genome_slide@region.names)
```

## WGS step 4: calculate Fst, pi, and Dxy for each window

This follows the same sequence as ddRAD: calculate $F_{ST}$, extract the window positions, count the SNPs in each window, and then add $pi$ and Dxy.

```{r wgs-fst-table}
wgs_positions <- as.numeric(wgs_vcf@fix[, "POS"])

wgs_genome_slide <- F_ST.stats(wgs_genome_slide, mode = "nucleotide")
wgs_fst_values <- as.data.frame(get.F_ST(wgs_genome_slide, mode = FALSE, pairwise = TRUE)[1])
names(wgs_fst_values) <- "fst"
wgs_fst_values$fst[wgs_fst_values$fst < 0] <- 0

wgs_fst_values$start_pos <- as.numeric(sub(" -.*", "", wgs_genome_slide@region.names))
wgs_fst_values$end_pos <- as.numeric(sub(" :", "", sub(".*- ", "", wgs_genome_slide@region.names)))
wgs_fst_values$midpoint_mb <- (wgs_fst_values$start_pos + wgs_fst_values$end_pos) / 2e6
wgs_fst_values$n_snps <- mapply(
  function(start, end) sum(wgs_positions >= start & wgs_positions <= end),
  wgs_fst_values$start_pos,
  wgs_fst_values$end_pos
)

wgs_genome_slide <- diversity.stats(wgs_genome_slide, pi = TRUE)
wgs_pi_values <- as.data.frame(wgs_genome_slide@Pi)
names(wgs_pi_values) <- c("pi_A2G", "pi_A2S")

wgs_genome_slide <- diversity.stats.between(wgs_genome_slide)
wgs_dxy_values <- as.data.frame(wgs_genome_slide@nuc.diversity.between)
names(wgs_dxy_values) <- "dxy"

wgs_stats <- cbind(wgs_fst_values, wgs_pi_values, wgs_dxy_values)
wgs_stats$dataset <- "WGS"
wgs_stats <- wgs_stats[, c("dataset", "start_pos", "end_pos", "midpoint_mb", "n_snps", "fst", "pi_A2G", "pi_A2S", "dxy")]
wgs_stats$pi_overall <- rowMeans(wgs_stats[, c("pi_A2G", "pi_A2S")], na.rm = TRUE)

wgs_smooth <- list(
  fst = predict(loess(fst ~ midpoint_mb, data = wgs_stats, span = 0.15, na.action = na.exclude)),
  pi_A2G = predict(loess(pi_A2G ~ midpoint_mb, data = wgs_stats, span = 0.15, na.action = na.exclude)),
  pi_A2S = predict(loess(pi_A2S ~ midpoint_mb, data = wgs_stats, span = 0.15, na.action = na.exclude)),
  pi_overall = predict(loess(pi_overall ~ midpoint_mb, data = wgs_stats, span = 0.15, na.action = na.exclude)),
  dxy = predict(loess(dxy ~ midpoint_mb, data = wgs_stats, span = 0.15, na.action = na.exclude))
)

wgs_fst <- wgs_stats[, c("dataset", "start_pos", "end_pos", "midpoint_mb", "n_snps", "fst")]

write.table(
  wgs_fst,
  file.path(output_dir, "wgs_scaffold1_window_fst.tsv"),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)

head(wgs_fst)
```

## WGS step 5: plot the WGS window scan

Now plot the WGS window summary to compare the positions of peaks and troughs along the same scaffold.

```{r wgs-plot, fig.width=9, fig.height=6}
wgs_plot_file <- file.path(output_dir, "wgs_scaffold1_fst.png")

png(wgs_plot_file, width = 9, height = 6, units = "in", res = 320)
plot(
  wgs_stats$midpoint_mb,
  wgs_stats$fst,
  ylim = c(0, 3.1),
  main = "Scaffold 1 - WGS",
  ylab = "FST",
  xlab = "Position on scaffold 1 (Mb)"
)
lines(wgs_stats$midpoint_mb, wgs_smooth$fst, col = plot_colors["fst"], lwd = 3)
lines(wgs_stats$midpoint_mb, wgs_smooth$pi_A2G, col = plot_colors["pi_A2G"], lwd = 3)
lines(wgs_stats$midpoint_mb, wgs_smooth$pi_A2S, col = plot_colors["pi_A2S"], lwd = 3)
lines(wgs_stats$midpoint_mb, wgs_smooth$pi_overall, col = plot_colors["pi_overall"], lwd = 3)
lines(wgs_stats$midpoint_mb, wgs_smooth$dxy, col = plot_colors["dxy"], lwd = 3)
legend(
  "topright",
  legend = plot_legend_labels,
  col = plot_colors,
  lwd = 3,
  bty = "n",
  cex = 0.75
)
dev.off()

knitr::include_graphics(wgs_plot_file)
```

