---
title: "Practical 3 Part 3: scaffold 1 window scan"
output: html_document
---

## Hide setup details

This first chunk only sets notebook-wide options. Students can leave it unchanged.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

Load the R packages used for window-based statistics, reshaping tables, and plotting.

```{r libraries}
library(cowplot)
library(dplyr)
library(ggplot2)
library(PopGenome)
library(tidyr)
library(vcfR)
```

## Set the practical root directory

Move to the scripts directory before reading inputs or writing outputs.

```{r working-directory}
setwd("/home/rstudio/work/Evolutionary-Genomics-UNIFR-2026-SBL.20034/practical3/scripts")
getwd()
```

## Set plot theme and window size

Define the shared plotting style, the output folder, and the sliding-window size used in this scaffold scan.

```{r options}
theme_set(
        theme_minimal(base_size = 11) +
                theme(
                        plot.title.position = "plot",
                        panel.grid.minor = element_blank(),
                        legend.position = "bottom"
                )
)

output_dir <- "../outputs/03-scaffold1-window-scan"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

window_width <- 10000
window_jump <- 1000
```

## Read the WGS scaffold data

Load the WGS VCF, identify the two focal populations, and create the PopGenome window object.

```{r wgs-inputs}
wgs_vcf <- read.vcfR(
        "../dataset-practical3/03-scaffold1-window-scan/wgs_vcf/scaffold1/A2_wgs_scaffold1.vcf",
        verbose = FALSE
)
wgs_names <- vcfR2genlight(wgs_vcf)$ind.names
wgs_populations <- list(
        A2G = grep("A2G", wgs_names, value = TRUE, fixed = TRUE),
        A2S = grep("A2S", wgs_names, value = TRUE, fixed = TRUE)
)

wgs_data <- readData("../dataset-practical3/03-scaffold1-window-scan/wgs_vcf/scaffold1", format = "VCF", include.unknown = TRUE)
wgs_windows <- sliding.window.transform(wgs_data, width = window_width, jump = window_jump, type = 2)
wgs_windows <- set.populations(wgs_windows, wgs_populations, diploid = TRUE)
```

## Calculate WGS Fst by window

Compute one Fst value per genomic window and extract the start and end coordinates for each window.

```{r wgs-fst-table}
wgs_windows <- F_ST.stats(wgs_windows, mode = "nucleotide")
wgs_fst <- as.data.frame(get.F_ST(wgs_windows, mode = FALSE, pairwise = TRUE)[1])
names(wgs_fst) <- "fst"
wgs_fst$fst[wgs_fst$fst < 0] <- 0

wgs_positions <- data.frame(
        start_pos = as.numeric(sub(" -.*", "", wgs_windows@region.names)),
        end_pos = as.numeric(sub(" :", "", sub(".*- ", "", wgs_windows@region.names)))
)

wgs_fst <- cbind(wgs_positions, wgs_fst)
wgs_fst$midpoint_mb <- (wgs_fst$start_pos + wgs_fst$end_pos) / 2e6

head(wgs_fst)
```

## Calculate WGS diversity statistics

Add nucleotide diversity, Dxy, and Tajima's D so the scaffold scan includes more than just Fst.

```{r wgs-diversity-table}
wgs_windows <- diversity.stats(wgs_windows, pi = TRUE)
wgs_windows <- diversity.stats.between(wgs_windows)
wgs_windows <- suppressWarnings(neutrality.stats(wgs_windows, FAST = TRUE))

wgs_pi <- as.data.frame(wgs_windows@Pi)
names(wgs_pi) <- c("pi_A2G", "pi_A2S")

wgs_dxy <- as.data.frame(wgs_windows@nuc.diversity.between)
names(wgs_dxy) <- "dxy_A2G_A2S"

wgs_tajima <- as.data.frame(wgs_windows@Tajima.D)
names(wgs_tajima) <- c("tajima_A2G", "tajima_A2S")

wgs_stats <- cbind(wgs_fst, wgs_pi, wgs_dxy, wgs_tajima)
write.table(wgs_stats, file = file.path(output_dir, "wgs_scaffold1_window_stats.tsv"), sep = "\t", quote = FALSE, row.names = FALSE)

head(wgs_stats)
```

## Pick highlighted WGS windows

Select the strongest WGS peaks so they can be emphasized in the plots.

```{r wgs-highlight-windows}
wgs_highlight_windows <- wgs_stats[order(-wgs_stats$fst, wgs_stats$midpoint_mb), ]
wgs_highlight_windows <- head(wgs_highlight_windows, 3)
wgs_highlight_windows$start_mb <- wgs_highlight_windows$start_pos / 1e6
wgs_highlight_windows$end_mb <- wgs_highlight_windows$end_pos / 1e6
wgs_highlight_windows$label_y <- wgs_highlight_windows$fst + seq(0.03, by = 0.03, length.out = nrow(wgs_highlight_windows))
wgs_highlight_windows$label <- paste0("Peak ", seq_len(nrow(wgs_highlight_windows)), "\n", sprintf("%.2f-%.2f Mb", wgs_highlight_windows$start_mb, wgs_highlight_windows$end_mb))

wgs_highlight_windows
```

## Reshape WGS tables for plotting

Convert the WGS pi and Tajima's D tables to long format so each population can be plotted separately.

```{r wgs-long-tables}
wgs_pi_long <- wgs_stats |>
        dplyr::select(midpoint_mb, pi_A2G, pi_A2S) |>
        dplyr::rename(A2G = pi_A2G, A2S = pi_A2S) |>
        tidyr::pivot_longer(cols = -midpoint_mb, names_to = "series", values_to = "value") |>
        dplyr::filter(is.finite(value))

wgs_tajima_long <- wgs_stats |>
        dplyr::select(midpoint_mb, tajima_A2G, tajima_A2S) |>
        dplyr::rename(A2G = tajima_A2G, A2S = tajima_A2S) |>
        tidyr::pivot_longer(cols = -midpoint_mb, names_to = "series", values_to = "value") |>
        dplyr::filter(is.finite(value))

wgs_pi_long$series <- factor(wgs_pi_long$series, levels = c("A2G", "A2S"))
wgs_tajima_long$series <- factor(wgs_tajima_long$series, levels = c("A2G", "A2S"))
```

## Plot WGS Fst

This plot shows how differentiation changes across scaffold 1 in the WGS dataset.

```{r wgs-fst-plot, fig.width=9, fig.height=4}
wgs_fst_plot <- ggplot(wgs_stats, aes(x = midpoint_mb, y = fst)) +
        geom_rect(
                data = wgs_highlight_windows,
                aes(xmin = start_mb, xmax = end_mb, ymin = -Inf, ymax = Inf),
                inherit.aes = FALSE,
                fill = "#4C78A8",
                alpha = 0.08
        ) +
        geom_point(size = 0.8, alpha = 0.45, colour = "#7A3E00") +
        geom_smooth(method = "loess", se = FALSE, span = 0.18, linewidth = 0.55, colour = "#7A3E00") +
        geom_point(
                data = wgs_highlight_windows,
                aes(x = midpoint_mb, y = fst),
                inherit.aes = FALSE,
                shape = 21,
                size = 2.2,
                fill = "white",
                colour = "#7A3E00",
                stroke = 0.55
        ) +
        geom_label(
                data = wgs_highlight_windows,
                aes(x = midpoint_mb, y = label_y, label = label),
                inherit.aes = FALSE,
                size = 2.5,
                label.size = 0.15,
                fill = "white"
        ) +
        labs(title = "WGS scaffold 1 differentiation", x = "Scaffold 1 position (Mb)", y = "Fst")

ggsave(file.path(output_dir, "wgs_scaffold1_fst.png"), wgs_fst_plot, width = 9, height = 4, dpi = 320)
wgs_fst_plot
```

## Plot WGS nucleotide diversity

These panels show within-population nucleotide diversity across the scaffold for A2G and A2S.

```{r wgs-pi-plot, fig.width=9, fig.height=5}
wgs_pi_plot <- ggplot(wgs_pi_long, aes(x = midpoint_mb, y = value, colour = series)) +
        geom_rect(
                data = wgs_highlight_windows,
                aes(xmin = start_mb, xmax = end_mb, ymin = -Inf, ymax = Inf),
                inherit.aes = FALSE,
                fill = "#4C78A8",
                alpha = 0.08
        ) +
        geom_point(size = 0.75, alpha = 0.35) +
        geom_smooth(method = "loess", se = FALSE, span = 0.2, linewidth = 0.55) +
        facet_wrap(~series, ncol = 1, scales = "free_y") +
        scale_colour_manual(values = c(A2G = "#1B9E77", A2S = "#D95F02")) +
        labs(title = "WGS nucleotide diversity", x = "Scaffold 1 position (Mb)", y = expression(pi), colour = NULL) +
        theme(legend.position = "none")

ggsave(file.path(output_dir, "wgs_scaffold1_pi.png"), wgs_pi_plot, width = 9, height = 5, dpi = 320)
wgs_pi_plot
```

## Plot WGS Dxy

This plot shows absolute divergence between A2G and A2S across scaffold 1 in the WGS dataset.

```{r wgs-dxy-plot, fig.width=9, fig.height=4}
wgs_dxy_plot <- ggplot(wgs_stats, aes(x = midpoint_mb, y = dxy_A2G_A2S)) +
        geom_rect(
                data = wgs_highlight_windows,
                aes(xmin = start_mb, xmax = end_mb, ymin = -Inf, ymax = Inf),
                inherit.aes = FALSE,
                fill = "#4C78A8",
                alpha = 0.08
        ) +
        geom_point(size = 0.8, alpha = 0.45, colour = "#B22222") +
        geom_smooth(method = "loess", se = FALSE, span = 0.18, linewidth = 0.55, colour = "#B22222") +
        labs(title = "WGS divergence", x = "Scaffold 1 position (Mb)", y = "Dxy")

ggsave(file.path(output_dir, "wgs_scaffold1_dxy.png"), wgs_dxy_plot, width = 9, height = 4, dpi = 320)
wgs_dxy_plot
```

## Plot WGS Tajima's D

These panels show Tajima's D across the scaffold for the two focal populations.

```{r wgs-tajima-plot, fig.width=9, fig.height=5}
wgs_tajima_plot <- ggplot(wgs_tajima_long, aes(x = midpoint_mb, y = value, colour = series)) +
        geom_rect(
                data = wgs_highlight_windows,
                aes(xmin = start_mb, xmax = end_mb, ymin = -Inf, ymax = Inf),
                inherit.aes = FALSE,
                fill = "#4C78A8",
                alpha = 0.08
        ) +
        geom_hline(yintercept = 0, linewidth = 0.35, colour = "grey55") +
        geom_point(size = 0.75, alpha = 0.35) +
        geom_smooth(method = "loess", se = FALSE, span = 0.2, linewidth = 0.55) +
        facet_wrap(~series, ncol = 1, scales = "free_y") +
        scale_colour_manual(values = c(A2G = "#1B9E77", A2S = "#D95F02")) +
        labs(title = "WGS Tajima's D", x = "Scaffold 1 position (Mb)", y = "Tajima's D", colour = NULL) +
        theme(legend.position = "none")

ggsave(file.path(output_dir, "wgs_scaffold1_tajima.png"), wgs_tajima_plot, width = 9, height = 5, dpi = 320)
wgs_tajima_plot
```

## Combine the WGS panels

Place the four WGS plots into one vertical summary panel for teaching and comparison.

```{r wgs-summary-plot, fig.width=10, fig.height=14}
wgs_summary_plot <- cowplot::plot_grid(wgs_fst_plot, wgs_pi_plot, wgs_dxy_plot, wgs_tajima_plot, ncol = 1, align = "v")

ggsave(file.path(output_dir, "wgs_scaffold1_summary.png"), wgs_summary_plot, width = 10, height = 14, dpi = 320)
wgs_summary_plot
```

## Read the ddRAD scaffold data

Load the ddRAD VCF, identify the two focal populations, and create the PopGenome window object.

```{r ddrad-inputs}
ddrad_vcf <- read.vcfR(
        "../dataset-practical3/03-scaffold1-window-scan/ddRAD_vcf/scaffold_1/A2_ddRAD_scaffold_1.vcf",
        verbose = FALSE
)
ddrad_names <- vcfR2genlight(ddrad_vcf)$ind.names
ddrad_populations <- list(
        A2G = grep("A2G", ddrad_names, value = TRUE, fixed = TRUE),
        A2S = grep("A2S", ddrad_names, value = TRUE, fixed = TRUE)
)

ddrad_data <- readData("../dataset-practical3/03-scaffold1-window-scan/ddRAD_vcf/scaffold_1", format = "VCF", include.unknown = TRUE)
ddrad_windows <- sliding.window.transform(ddrad_data, width = window_width, jump = window_jump, type = 2)
ddrad_windows <- set.populations(ddrad_windows, ddrad_populations, diploid = TRUE)
```

## Calculate ddRAD Fst by window

Compute one Fst value per genomic window and extract the start and end coordinates for each ddRAD window.

```{r ddrad-fst-table}
ddrad_windows <- F_ST.stats(ddrad_windows, mode = "nucleotide")
ddrad_fst <- as.data.frame(get.F_ST(ddrad_windows, mode = FALSE, pairwise = TRUE)[1])
names(ddrad_fst) <- "fst"
ddrad_fst$fst[ddrad_fst$fst < 0] <- 0

ddrad_positions <- data.frame(
        start_pos = as.numeric(sub(" -.*", "", ddrad_windows@region.names)),
        end_pos = as.numeric(sub(" :", "", sub(".*- ", "", ddrad_windows@region.names)))
)

ddrad_fst <- cbind(ddrad_positions, ddrad_fst)
ddrad_fst$midpoint_mb <- (ddrad_fst$start_pos + ddrad_fst$end_pos) / 2e6

head(ddrad_fst)
```

## Calculate ddRAD diversity statistics

Add nucleotide diversity, Dxy, and Tajima's D so the ddRAD scaffold scan can be compared directly to the WGS scan.

```{r ddrad-diversity-table}
ddrad_windows <- diversity.stats(ddrad_windows, pi = TRUE)
ddrad_windows <- diversity.stats.between(ddrad_windows)
ddrad_windows <- suppressWarnings(neutrality.stats(ddrad_windows, FAST = TRUE))

ddrad_pi <- as.data.frame(ddrad_windows@Pi)
names(ddrad_pi) <- c("pi_A2G", "pi_A2S")

ddrad_dxy <- as.data.frame(ddrad_windows@nuc.diversity.between)
names(ddrad_dxy) <- "dxy_A2G_A2S"

ddrad_tajima <- as.data.frame(ddrad_windows@Tajima.D)
names(ddrad_tajima) <- c("tajima_A2G", "tajima_A2S")

ddrad_stats <- cbind(ddrad_fst, ddrad_pi, ddrad_dxy, ddrad_tajima)
write.table(ddrad_stats, file = file.path(output_dir, "ddrad_scaffold1_window_stats.tsv"), sep = "\t", quote = FALSE, row.names = FALSE)

head(ddrad_stats)
```

## Pick highlighted ddRAD windows

Select the strongest ddRAD peaks so they can be emphasized in the plots.

```{r ddrad-highlight-windows}
ddrad_highlight_windows <- ddrad_stats[order(-ddrad_stats$fst, ddrad_stats$midpoint_mb), ]
ddrad_highlight_windows <- head(ddrad_highlight_windows, 3)
ddrad_highlight_windows$start_mb <- ddrad_highlight_windows$start_pos / 1e6
ddrad_highlight_windows$end_mb <- ddrad_highlight_windows$end_pos / 1e6
ddrad_highlight_windows$label_y <- ddrad_highlight_windows$fst + seq(0.03, by = 0.03, length.out = nrow(ddrad_highlight_windows))
ddrad_highlight_windows$label <- paste0("Peak ", seq_len(nrow(ddrad_highlight_windows)), "\n", sprintf("%.2f-%.2f Mb", ddrad_highlight_windows$start_mb, ddrad_highlight_windows$end_mb))

ddrad_highlight_windows
```

## Reshape ddRAD tables for plotting

Convert the ddRAD pi and Tajima's D tables to long format so each population can be plotted separately.

```{r ddrad-long-tables}
ddrad_pi_long <- ddrad_stats |>
        dplyr::select(midpoint_mb, pi_A2G, pi_A2S) |>
        dplyr::rename(A2G = pi_A2G, A2S = pi_A2S) |>
        tidyr::pivot_longer(cols = -midpoint_mb, names_to = "series", values_to = "value") |>
        dplyr::filter(is.finite(value))

ddrad_tajima_long <- ddrad_stats |>
        dplyr::select(midpoint_mb, tajima_A2G, tajima_A2S) |>
        dplyr::rename(A2G = tajima_A2G, A2S = tajima_A2S) |>
        tidyr::pivot_longer(cols = -midpoint_mb, names_to = "series", values_to = "value") |>
        dplyr::filter(is.finite(value))

ddrad_pi_long$series <- factor(ddrad_pi_long$series, levels = c("A2G", "A2S"))
ddrad_tajima_long$series <- factor(ddrad_tajima_long$series, levels = c("A2G", "A2S"))
```

## Plot ddRAD Fst

This plot shows how differentiation changes across scaffold 1 in the ddRAD dataset.

```{r ddrad-fst-plot, fig.width=9, fig.height=4}
ddrad_fst_plot <- ggplot(ddrad_stats, aes(x = midpoint_mb, y = fst)) +
        geom_rect(
                data = ddrad_highlight_windows,
                aes(xmin = start_mb, xmax = end_mb, ymin = -Inf, ymax = Inf),
                inherit.aes = FALSE,
                fill = "#F4A261",
                alpha = 0.1
        ) +
        geom_point(size = 0.8, alpha = 0.45, colour = "#7A3E00") +
        geom_smooth(method = "loess", se = FALSE, span = 0.18, linewidth = 0.55, colour = "#7A3E00") +
        geom_point(
                data = ddrad_highlight_windows,
                aes(x = midpoint_mb, y = fst),
                inherit.aes = FALSE,
                shape = 21,
                size = 2.2,
                fill = "white",
                colour = "#7A3E00",
                stroke = 0.55
        ) +
        geom_label(
                data = ddrad_highlight_windows,
                aes(x = midpoint_mb, y = label_y, label = label),
                inherit.aes = FALSE,
                size = 2.5,
                label.size = 0.15,
                fill = "white"
        ) +
        labs(title = "ddRAD scaffold 1 differentiation", x = "Scaffold 1 position (Mb)", y = "Fst")

ggsave(file.path(output_dir, "ddrad_scaffold1_fst.png"), ddrad_fst_plot, width = 9, height = 4, dpi = 320)
ddrad_fst_plot
```

## Plot ddRAD nucleotide diversity

These panels show within-population nucleotide diversity across the scaffold for A2G and A2S in the ddRAD dataset.

```{r ddrad-pi-plot, fig.width=9, fig.height=5}
ddrad_pi_plot <- ggplot(ddrad_pi_long, aes(x = midpoint_mb, y = value, colour = series)) +
        geom_rect(
                data = ddrad_highlight_windows,
                aes(xmin = start_mb, xmax = end_mb, ymin = -Inf, ymax = Inf),
                inherit.aes = FALSE,
                fill = "#F4A261",
                alpha = 0.1
        ) +
        geom_point(size = 0.75, alpha = 0.35) +
        geom_smooth(method = "loess", se = FALSE, span = 0.2, linewidth = 0.55) +
        facet_wrap(~series, ncol = 1, scales = "free_y") +
        scale_colour_manual(values = c(A2G = "#1B9E77", A2S = "#D95F02")) +
        labs(title = "ddRAD nucleotide diversity", x = "Scaffold 1 position (Mb)", y = expression(pi), colour = NULL) +
        theme(legend.position = "none")

ggsave(file.path(output_dir, "ddrad_scaffold1_pi.png"), ddrad_pi_plot, width = 9, height = 5, dpi = 320)
ddrad_pi_plot
```

## Plot ddRAD Dxy

This plot shows absolute divergence between A2G and A2S across scaffold 1 in the ddRAD dataset.

```{r ddrad-dxy-plot, fig.width=9, fig.height=4}
ddrad_dxy_plot <- ggplot(ddrad_stats, aes(x = midpoint_mb, y = dxy_A2G_A2S)) +
        geom_rect(
                data = ddrad_highlight_windows,
                aes(xmin = start_mb, xmax = end_mb, ymin = -Inf, ymax = Inf),
                inherit.aes = FALSE,
                fill = "#F4A261",
                alpha = 0.1
        ) +
        geom_point(size = 0.8, alpha = 0.45, colour = "#B22222") +
        geom_smooth(method = "loess", se = FALSE, span = 0.18, linewidth = 0.55, colour = "#B22222") +
        labs(title = "ddRAD divergence", x = "Scaffold 1 position (Mb)", y = "Dxy")

ggsave(file.path(output_dir, "ddrad_scaffold1_dxy.png"), ddrad_dxy_plot, width = 9, height = 4, dpi = 320)
ddrad_dxy_plot
```

## Plot ddRAD Tajima's D

These panels show Tajima's D across the scaffold for the two focal populations in the ddRAD dataset.

```{r ddrad-tajima-plot, fig.width=9, fig.height=5}
ddrad_tajima_plot <- ggplot(ddrad_tajima_long, aes(x = midpoint_mb, y = value, colour = series)) +
        geom_rect(
                data = ddrad_highlight_windows,
                aes(xmin = start_mb, xmax = end_mb, ymin = -Inf, ymax = Inf),
                inherit.aes = FALSE,
                fill = "#F4A261",
                alpha = 0.1
        ) +
        geom_hline(yintercept = 0, linewidth = 0.35, colour = "grey55") +
        geom_point(size = 0.75, alpha = 0.35) +
        geom_smooth(method = "loess", se = FALSE, span = 0.2, linewidth = 0.55) +
        facet_wrap(~series, ncol = 1, scales = "free_y") +
        scale_colour_manual(values = c(A2G = "#1B9E77", A2S = "#D95F02")) +
        labs(title = "ddRAD Tajima's D", x = "Scaffold 1 position (Mb)", y = "Tajima's D", colour = NULL) +
        theme(legend.position = "none")

ggsave(file.path(output_dir, "ddrad_scaffold1_tajima.png"), ddrad_tajima_plot, width = 9, height = 5, dpi = 320)
ddrad_tajima_plot
```

## Combine the ddRAD panels

Place the four ddRAD plots into one vertical summary panel for teaching and comparison.

```{r ddrad-summary-plot, fig.width=10, fig.height=14}
ddrad_summary_plot <- cowplot::plot_grid(ddrad_fst_plot, ddrad_pi_plot, ddrad_dxy_plot, ddrad_tajima_plot, ncol = 1, align = "v")

ggsave(file.path(output_dir, "ddrad_scaffold1_summary.png"), ddrad_summary_plot, width = 10, height = 14, dpi = 320)
ddrad_summary_plot
```

## Compare ddRAD and WGS Fst

Finish with a simple two-panel comparison so students can inspect how the ddRAD and WGS Fst scans line up.

```{r ddrad-vs-wgs-fst, fig.width=9, fig.height=7}
comparison_plot <- cowplot::plot_grid(ddrad_fst_plot, wgs_fst_plot, ncol = 1, align = "v")

ggsave(file.path(output_dir, "ddrad_vs_wgs_fst.png"), comparison_plot, width = 9, height = 7, dpi = 320)
comparison_plot
```
