---
title: "Setup – Practical 3 Tools"
subtitle: "structure · CLUMPP · structureHarvester"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

## Overview

This notebook installs everything you need for **Practical 3**:

| Tool | Type | Purpose |
|------|------|---------|
| `structure` | CLI (Bioconda) | Bayesian clustering of multilocus genotypes |
| `CLUMPP` | CLI (binary) | Aligns cluster memberships across multiple STRUCTURE runs |
| `structureHarvester` | CLI (Python) | Extracts Δ*K* and best-*K* statistics from STRUCTURE output |

All three tools are installed into `<repo_root>/tools/bin/` so they can be
called from any notebook or terminal with a single `source` command.

> **Tip:** Run this notebook **once** from the repository root in RStudio.
> After that the tools are persistent – you only need to re-source the env
> script at the start of each new session.

---

## 1  Locate the repository root

The chunk below detects the repo root automatically.  If RStudio opened a
*sub-folder* project instead of the workspace root, set `repo_root` manually
before running anything else.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, comment = "")
```

```{r repo-root}
find_repo_root <- function(start = getwd()) {
  path <- normalizePath(start, mustWork = TRUE)
  for (i in seq_len(10)) {
    if (file.exists(file.path(path, "setup", "setup_practical3_tools.sh"))) {
      return(path)
    }
    parent <- dirname(path)
    if (parent == path) break
    path <- parent
  }
  "/home/rstudio/work/Evolutionary-Genomics-UNIFR-2026-SBL.20034"
}

repo_root <- find_repo_root()
tools_dir <- file.path(repo_root, "tools")
bin_dir   <- file.path(tools_dir, "bin")
env3_dir  <- file.path(tools_dir, "env3")
installer <- file.path(repo_root, "setup", "setup_practical3_tools.sh")

cat("repo_root :", repo_root, "\n")
cat("tools_dir :", tools_dir, "\n")
cat("installer :", installer, "\n")
stopifnot("Installer script not found – check repo_root above" = file.exists(installer))
```

---

## 2  Run the shell installer

The installer (`setup_practical3_tools.sh`) is **idempotent** – re-running it
is safe and will skip steps that have already been completed.

It will:

1. Download **micromamba** (a tiny conda package manager) into `tools/.micromamba/`
2. Create a local conda environment `tools/env3/` with **structure** (Bioconda) and Python
3. Download the **CLUMPP** pre-compiled Linux binary
4. Clone **structureHarvester** from GitHub
5. Write wrapper scripts into `tools/bin/`
6. Generate `tools/env_practical3.sh`
7. Add PATH hooks to `~/.bashrc`, `~/.profile`, and `~/.Renviron`

```{r run-installer, eval=FALSE}
stopifnot(file.exists(installer))
ret <- system2("bash", installer)
if (ret != 0L) stop("Installer exited with code ", ret, ". Check the output above.")
cat("\nInstaller finished successfully.\n")
```

---

## 3  Activate tools in the current R session

After installation the tools are on disk, but the current R session still
needs to know where to find them.  Run this chunk once per session.

```{r activate-session, eval=FALSE}
Sys.setenv(
  PATH = paste(bin_dir,
               file.path(env3_dir, "bin"),
               Sys.getenv("PATH"),
               sep = ":"),
  LD_LIBRARY_PATH = paste(file.path(env3_dir, "lib"),
                          Sys.getenv("LD_LIBRARY_PATH"),
                          sep = ":")
)
cat("PATH updated for this R session.\n")
```

---

## 4  Verify the installation

Run these checks to confirm every tool is accessible.

```{r verify-which, eval=FALSE}
tools_to_check <- c("structure", "CLUMPP", "structureHarvester")

cat("--- Tool locations ---\n")
for (t in tools_to_check) {
  path <- Sys.which(t)
  if (nchar(path) > 0) {
    cat(sprintf("  %-25s  %s\n", t, path))
  } else {
    cat(sprintf("  %-25s  NOT FOUND\n", t))
  }
}
```

```{r verify-structure, eval=FALSE}
cat("\n--- structure (first line of help) ---\n")
system("structure --help 2>&1 | head -3")
```

```{r verify-clumpp, eval=FALSE}
cat("\n--- CLUMPP ---\n")
system("CLUMPP --help 2>&1 | head -3")
```

```{r verify-structureharvester, eval=FALSE}
cat("\n--- structureHarvester ---\n")
system("structureHarvester --help 2>&1 | head -5")
```

```{r inspect-tools-dir, eval=FALSE}
cat("\n--- tools/bin contents ---\n")
cat(paste(sort(list.files(bin_dir)), collapse = "\n"), "\n")
```

---

## 5  Quick-start for subsequent sessions

At the top of every Practical 3 notebook, add the following activation chunk:

```r
# Activate Practical 3 tools
tools_dir <- file.path(
  "/home/rstudio/work/Evolutionary-Genomics-UNIFR-2026-SBL.20034", "tools")
Sys.setenv(
  PATH = paste(file.path(tools_dir, "bin"),
               file.path(tools_dir, "env3", "bin"),
               Sys.getenv("PATH"), sep = ":"),
  LD_LIBRARY_PATH = paste(file.path(tools_dir, "env3", "lib"),
                          Sys.getenv("LD_LIBRARY_PATH"), sep = ":")
)
```

Or, from the terminal / a `system()` call:

```bash
source tools/env_practical3.sh
```

---

## Notes on the tools

### structure

`structure` is installed from the Bioconda channel into the local conda
environment `tools/env3`.  The wrapper at `tools/bin/structure` forwards all
arguments to `tools/env3/bin/structure`.

### CLUMPP

CLUMPP is distributed as a pre-compiled 64-bit Linux binary by the Rosenberg
lab.  It is placed directly in `tools/bin/CLUMPP`.

### structureHarvester

structureHarvester is a Python script cloned from
[dentearl/structureHarvester](https://github.com/dentearl/structureHarvester).
The wrapper at `tools/bin/structureHarvester` calls the script with the Python
interpreter from `tools/env3`.