---
title: "Setup – Practical 2 Tools"
subtitle: "car · poolfstat · samtools · bwa · java · npstat · popoolation2"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

## Overview

This notebook installs everything you need for **Practical 2**:

| Tool | Type | Purpose |
|------|------|---------|
| `car` | R package | Companion to Applied Regression – linear models / ANOVA |
| `poolfstat` | R package | Pool-seq FST and differentiation statistics |
| `samtools` | CLI (conda) | SAM/BAM sorting, indexing, and manipulation |
| `bwa` | CLI (conda) | Short-read alignment (BWA-MEM) |
| `java` | CLI (conda, OpenJDK) | Java runtime – required by popoolation2 `.jar` utilities |
| `npstat` | CLI (compiled C) | Summary statistics from pool-seq data |
| `popoolation2` | CLI (Perl) | Sliding-window FST on Pool-seq data |

The CLI tools (`npstat` and `popoolation2`) are installed into
`<repo_root>/tools/bin/` and a small convenience script
`tools/env_practical2.sh` is generated so you can activate them any time
by running `source tools/env_practical2.sh`.

> **Tip:** Run this notebook **once** from the repository root in RStudio.
> After that the tools are persistent – you only need to re-source the env
> script at the start of each new session.

---

## 1  Locate the repository root

The chunk below detects the repo root automatically.  If RStudio opened a
*sub-folder* project instead of the workspace root, set `repo_root` manually
before running anything else.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, comment = "")
```

```{r repo-root}
# Auto-detect: walk upward from the current working directory until we find
# the install script.  Falls back to the hard-coded Renku path.
find_repo_root <- function(start = getwd()) {
  path <- normalizePath(start, mustWork = TRUE)
  for (i in seq_len(10)) {
    if (file.exists(file.path(path, "setup", "setup_practical2_tools.sh"))) {
      return(path)
    }
    parent <- dirname(path)
    if (parent == path) break
    path <- parent
  }
  # Fallback for Renku / standard deployment
  "/home/rstudio/work/Evolutionary-Genomics-UNIFR-2026-SBL.20034"
}

repo_root  <- find_repo_root()
tools_dir  <- file.path(repo_root, "tools")
bin_dir    <- file.path(tools_dir, "bin")
env2_dir   <- file.path(tools_dir, "env2")
perl5_dir  <- file.path(tools_dir, "perl5", "lib", "perl5")
installer  <- file.path(repo_root, "setup", "setup_practical2_tools.sh")

cat("repo_root :", repo_root, "\n")
cat("tools_dir :", tools_dir, "\n")
cat("installer :", installer, "\n")
stopifnot("Installer script not found – check repo_root above" = file.exists(installer))
```

---

## 2  Run the shell installer

The installer (`setup_practical2_tools.sh`) is **idempotent** – re-running it
is safe and will skip steps that have already been completed.

It will:

1. Download **micromamba** (a tiny conda package manager) into `tools/.micromamba/`
2. Create a local conda environment `tools/env2/` with **samtools**, **bwa**, **OpenJDK**, GSL, Perl, and **perl-doc**
3. Clone and **compile** `npstat`
4. Download and unpack `popoolation2`
5. Write wrapper scripts into `tools/bin/`
6. Generate `tools/env_practical2.sh`
7. Add PATH hooks to `~/.bashrc`, `~/.profile`, and `~/.Renviron`

```{r run-installer, eval=FALSE}
stopifnot(file.exists(installer))
ret <- system2("bash", installer)
if (ret != 0L) stop("Installer exited with code ", ret, ". Check the output above.")
cat("\nInstaller finished successfully.\n")
```

---

## 3  Activate tools in the current R session

After installation the tools are on disk, but the current R session still
needs to know where to find them.  Run this chunk once per session.

```{r activate-session, eval=FALSE}
Sys.setenv(
  PATH = paste(bin_dir,
               file.path(env2_dir, "bin"),
               Sys.getenv("PATH"),
               sep = ":"),
  LD_LIBRARY_PATH = paste(file.path(env2_dir, "lib"),
                          Sys.getenv("LD_LIBRARY_PATH"),
                          sep = ":"),
  PERL5LIB = paste(perl5_dir,
                   Sys.getenv("PERL5LIB"),
                   sep = ":")
)
cat("PATH updated for this R session.\n")
```

---

## 4  Install R packages

The R packages are installed into your project's renv library so that they
are available in every subsequent session without re-running this step.

```{r install-r-packages, eval=FALSE}
r_pkgs <- c(
  # Core packages for Practical 2
  "car",          # linear models / ANOVA companion
  "poolfstat",    # Pool-seq FST statistics
  # Supporting packages commonly used alongside the above
  "vcfR",         # read/manipulate VCF files
  "tidyverse",    # data wrangling and plotting
  "data.table",   # fast tabular data
  "locfit"        # local regression (used by some pool-seq analyses)
)

missing_pkgs <- r_pkgs[!vapply(r_pkgs, requireNamespace, logical(1), quietly = TRUE)]

if (length(missing_pkgs) > 0) {
  message("Installing missing R packages: ", paste(missing_pkgs, collapse = ", "))
  install.packages(missing_pkgs, repos = "https://cloud.r-project.org")
} else {
  message("All R packages already installed.")
}
```

---

## 5  Verify the installation

Run these checks to confirm every tool is accessible from the R session.

```{r verify-cli, eval=FALSE}
tools_to_check <- c(
  "samtools",
  "bwa",
  "java",
  "npstat",
  "fst-sliding.pl",
  "Variance-sliding.pl",
  "fisher-test.pl"
)

cat("--- CLI tools ---\n")
for (t in tools_to_check) {
  path <- Sys.which(t)
  if (nchar(path) > 0) {
    cat(sprintf("  %-30s  %s\n", t, path))
  } else {
    cat(sprintf("  %-30s  NOT FOUND\n", t))
  }
}
```

```{r verify-samtools, eval=FALSE}
cat("\n--- samtools version ---\n")
system("samtools --version 2>&1 | head -2")
```

```{r verify-bwa, eval=FALSE}
cat("\n--- bwa version ---\n")
system("bwa 2>&1 | head -3")
```

```{r verify-java, eval=FALSE}
cat("\n--- java version ---\n")
system("java -version 2>&1")
```

```{r verify-npstat, eval=FALSE}
cat("\n--- npstat version ---\n")
system("npstat --help 2>&1 | head -5")
```

```{r verify-popoolation2, eval=FALSE}
cat("\n--- popoolation2 (fst-sliding.pl) ---\n")
system("fst-sliding.pl --help 2>&1 | head -5")
```

```{r verify-r-packages, eval=FALSE}
cat("\n--- R packages ---\n")
for (pkg in c("car", "poolfstat")) {
  installed <- requireNamespace(pkg, quietly = TRUE)
  cat(sprintf("  %-15s  %s\n", pkg, if (installed) "OK" else "MISSING"))
}
```

```{r inspect-tools-dir, eval=FALSE}
cat("\n--- tools/bin contents ---\n")
cat(paste(sort(list.files(bin_dir)), collapse = "\n"), "\n")
```

---

## 6  Quick-start for subsequent sessions

At the top of every Practical 2 notebook, add the following activation chunk
(it is safe to run even if the environment variable was already set):

```r
# Activate Practical 2 tools
tools_dir <- file.path(
  "/home/rstudio/work/Evolutionary-Genomics-UNIFR-2026-SBL.20034", "tools")
Sys.setenv(
  PATH = paste(file.path(tools_dir, "bin"),
               file.path(tools_dir, "env2", "bin"),
               Sys.getenv("PATH"), sep = ":"),
  LD_LIBRARY_PATH = paste(file.path(tools_dir, "env2", "lib"),
                          Sys.getenv("LD_LIBRARY_PATH"), sep = ":"),
  PERL5LIB = paste(file.path(tools_dir, "perl5", "lib", "perl5"),
                   Sys.getenv("PERL5LIB"), sep = ":")
)
```

Or, from the terminal / a `system()` call:

```bash
source tools/env_practical2.sh
```
